---
title: "Genomic annotation in Bioconductor: The general situation"
author: "Vince"
date: "March 19, 2015"
output: html_document
layout: page
toc: yes
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(AnnotationDbi)
library(BSgenome.Hsapiens.NCBI.GRCh38)
library(BSgenome.Hsapiens.UCSC.hg19)
library(Biostrings)
library(GenomicRanges)
library(IRanges)
library(Homo.sapiens)
library(grid)
library(png)
library(KEGGREST)
library(GO.db)
library(org.Hs.eg.db)
})
```

# Basic annotation resources and their discovery

In this document we will review Bioconductor's facilities for
handling and annotating genomic sequence.  We'll look at
reference genomic sequence, transcripts and genes, and
conclude with gene pathways.  Keep in mind that our ultimate aim
is to use annotation information to help produce reliable
interpretations of genomic experiments.  A basic objective of
Bioconductor is to make it easy to incorporate
information on genome structure and function 
into statistical analysis procedures.

## A simple hierarchy of annotation concepts

Bioconductor includes many different types of genomic annotation.
We can think of these annotation resources in a hierarchical structure.

- At the base is the reference genomic sequence for an organism.
This is always arranged into chromosomes, specified by linear
sequences of nucleotides.

- Above this is the organization of chromosomal sequence into
regions of interest.  The most prominent regions of interest are
genes, but other structures like SNPs or CpG sites are
annotated as well.  Genes have internal structure,
with parts that are transcribed and parts that are not,
and "gene models" define the ways in which
these structures are labeled and laid out in genomic coordinates.

- Above this is the organization of genes or gene products into
groups with shared structural or functional properties.  Examples
include pathways, groups of genes found together in cells, or
identified as cooperating in biological processes.

## Discovering available reference genomes

Bioconductor's collection of annotation packages brings
all elements of this hierarchy into a programmable environment.
Reference genomic sequences are managed using the infrastructure
of the Biostrings and BSgenome packages, and the `available.genomes`
function lists the reference genome build for humans and
various model organisms now available.

```{r lkbs}
library(Biostrings)
ag = available.genomes()
length(ag)
head(ag)
```

## Reference build versions are important

The reference build for an organism is created de novo
and then refined as algorithms and sequenced data improve.
For humans, the Genome Research Consortium signed off on
build 37 in 2009, and on build 38 in 2013.

Once a reference build is completed, it becomes easy to
perform informative genomic sequence analysis on individuals, because one can
focus on regions that are known to harbor allelic diversity.

Note that the genome sequence packages have long names
that include build versions.  It is very important to avoid
mixing coordinates from different reference builds.
In the liftOver video we show how to convert genomic coordinates of
features between different reference builds, using the UCSC
"liftOver" utility interfaced to R in the `rtracklayer` package.

To help users avoid mixing up data collected on incompatible
genomic coordinate systems from different reference builds, we
include a "genome" tag that can be filled out for most objects
that hold sequence information.  We'll see some examples of
this shortly.  Software for sequence comparison can check
for compatible tags on the sequences
being compared, and thereby help to ensure meaningful results.

# A reference genomic sequence for H. sapiens

The reference sequence for *Homo sapiens* is acquired by installing
and attaching
a single package.  This is in contrast to downloading and parsing
FASTA files.  The package defines an object `Hsapiens`
that is the source of chromosomal sequence, but when
evaluated on its own
provides a report of the origins of the sequence data that
it contains.

```{r lkhs}
library(BSgenome.Hsapiens.UCSC.hg19)
Hsapiens
head(genome(Hsapiens))  # see the tag
```

We acquire a chromosome's sequence using the `$` operator.
```{r lkc}
Hsapiens$chr17
```

# The transcripts and genes for a reference sequence

The `TxDb` family of packages and data objects manages
information on transcripts and gene models.

```{r gettx}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb = TxDb.Hsapiens.UCSC.hg19.knownGene # abbreviate
txdb
```

We can use `genes()` to get the addresses of genes using 
Entrez Gene IDs.

```{r getg}
ghs = genes(txdb)
ghs
```

# Your data will be someone else's annotation: import/export

The ENCODE project is a good example of the idea that today's
experiment is tomorrow's annotation.  You should think of your
own experiments in the same way.  (Of course, for an experiment
to serve as reliable and durable annotation it must address an
important question about genomic structure or function and must
answer it with an appropriate, properly executed protocol.
ENCODE is noteworthy for linking the protocols to the data
very explicitly.)

What we have to watch out for is the idea that annotation is somehow
permanently correct, isolated from the cacophony of research progress
at the boundaries of knowledge.  We have seen that even the
reference sequences of human chromosomes are subject to revision.
We have treated, in our use of the ERBS package, results of experiments
that we don't know too much about, as defining ER binding sites for
potential biological interpretation.  The uncertainty, variable
quality of peak identification, has not 
been explicitly reckoned but it should be.

Bioconductor has taken pains to acknowledge many facets of this situation.
We maintain archives of prior versions of software
and annotation so that past work can be checked
or revised.  We update central annotation resources twice a year so that
there is stability for ongoing work as well as access to new knowledge.
And we have made it simple to import and to create representations
of experimental and annotation data.

As an example, return to the ER binding data.  These were published
by ENCODE as narrowPeak files.  This is ascii text at its base, so
can be imported as a set of textual lines with no difficulty.
If there is sufficient regularity to the record fields, 
the file could be imported as a table.

We want to go beyond this, so that the import is usable as a computable
object as rapidly as possible.  Recognizing the connection between the
narrowPeak and bedGraph formats, we can import immediately to a
GRanges.

To illustrate this, we find the path to the narrowPeak raw data
file in the ERBS package.
```{r getgraa}
f1 = dir(system.file("extdata",package="ERBS"), full=TRUE)[1]
readLines(f1, 4) # look at a few lines
library(rtracklayer)
imp = import(f1, format="bedGraph")
imp
genome(imp)  # genome identifier tag not set, but you should set it
```

We have a GRanges in one stroke.  There are some additional fields
in the metadata columns that will need to be specified, but if we
are interested only in the ranges, we are done, with the exception 
of adding the genome metadata to protect against illegitimate
combination with data recorded in an incompatible coordinate system.

For communicating with other scientists or systems we have two
main options.  First, we can save the GRanges as an "RData" object,
easily transmitted to another R user for immediate use.  Second,
we can export in another standard format.  For example, if we
are interested only in interval addresses and the binding scores,
it is sufficient to save in "bed" format.

```{r doexp}
export(imp, "demoex.bed")  # implicit format choice
cat(readLines("demoex.bed", n=5), sep="\n")
```

We have carried out a "round trip" of importing, modeling, and exporting
experimental data that can be integrated with other data to advance
biological understanding.

To conclude this group, I mention a newly developed package
called AnnotationHub that can be used to obtain GRanges or other
suitably designed containers for institutionally curated annotation.
Here we will show that there are a number of experimental data
objects related to the HepG2 cell line available through AnnotationHub.

```{r lkah}
library(AnnotationHub)
ah = AnnotationHub()
ah
query(ah, "HepG2")
```

Advanced users will profit from getting acquainted with this package.

# The NCBI Entrez Gene annotation maps

Packages named org.*.eg.db collect information at the gene level
with links to location, protein product identifiers, KEGG pathway and
GO terms, PMIDs of papers mentioning genes, and to
identifiers for other annotation resources.

```{r lkor}
library(org.Hs.eg.db)
keytypes(org.Hs.eg.db) # columns() gives same answer
```
```{r lkt}
head(select(org.Hs.eg.db, keys="ORMDL3", keytype="SYMBOL", 
   columns="PMID"))
```

# Resources for gene sets and pathways

## Gene Ontology

[Gene Ontology](http://www.geneontology.org) (GO) is
a widely used structured vocabulary that organizes terms relevant to
the roles of genes and gene products in
- biological processes,
- molecular functions, and
- cellular components.
The vocabulary itself is intended to be relevant for all organisms.
It takes the form of a directed acyclic graph, with terms as
nodes and 'is-a' and 'part-of' relationships comprising
most of the links.

The *annotation* that links organism-specific genes to terms
in gene ontology is separate from the vocabulary itself, and
involves different types of evidence.  These are recorded
in Bioconductor annotation packages.

We have immediate access to the GO vocabulary
with the `GO.db` package.
```{r lkgo1}
library(GO.db)
GO.db # metadata
```

The keys/columns/select functionality of
`AnnotationDbi` is easy to use for mappings between ids,
terms and definitions.

```{r lkgo2}
k5 = keys(GO.db)[1:5]
cgo = columns(GO.db)
select(GO.db, keys=k5, columns=cgo[1:3])
```

The graphical structure of the vocabulary is encoded in
tables in a SQLite database.  We can query this using
the `RSQLite` interface.
```{r lkgo3}
con = GO_dbconn()
dbListTables(con)
```

The following query reveals some internal identifiers:
```{r lkids}
dbGetQuery(con, "select _id, go_id, term from go_term limit 5")
```
We can trace the `mitochondrion inheritance` term to
parent and grandparent terms:

```{r lkpath}
dbGetQuery(con, "select * from go_bp_parents where _id=27")
dbGetQuery(con, "select _id, go_id, term from go_term where _id=26638")
dbGetQuery(con, "select * from go_bp_parents where _id=26638")
dbGetQuery(con, "select _id, go_id, term from go_term where _id=5938")
```
It makes sense to regard "mitochondrion inheritance" as a conceptual
refinement of processes "mitochondrion distribution", and "organelle inheritance",
the two terms that are regarded as parents in this database scheme.

The entire database schema can be viewed with `GO_dbschema()`.

## KEGG: Kyoto Encyclopedia of Genes and Genomes

The KEGG annotation system has been available in Bioconductor 
since the latter's inception, but licensing of the database
has changed.  When we attach `KEGG.db` we see

```
> library(KEGG.db)
KEGG.db contains mappings based on older data because the original
  resource was removed from the the public domain before the most
  recent update was produced. This package should now be considered
  deprecated and future versions of Bioconductor may not have it
  available.  Users who want more current data are encouraged to look
  at the KEGGREST or reactome.db packages
```

Therefore we focus on KEGGREST, which requires active internet connection.
A very useful query resolution facility is based on Entrez identifiers.
The Entrez ID for BRCA2 is 675.  We'll perform a general query.
```{r dokr}
library(KEGGREST)
brca2K = keggGet("hsa:675")
names(brca2K[[1]])
```

The list of genes making up a pathway model 
can be obtained with another keggGet:
```{r dolkp}
brpat = keggGet("path:hsa05212")
names(brpat[[1]])
brpat[[1]]$GENE[seq(1,132,2)] # entrez gene ids
```

There is much to explore, and the KEGGREST package vignette provides
examples.  As a last illustration, we can acquire a static image of
the (human) pancreatic cancer pathway in which BRCA2 is implicated.

```{r getp,fig=TRUE}
library(png)
library(grid)
brpng = keggGet("hsa05212", "image")
grid.raster(brpng)
```



# A unified, self-describing approach

The OrganismDb packages simplify access to annotation.
Queries that succeed against TxDb, and org.[Nn].eg.db
can be directed at the OrganismDb object.

```{r getl}
library(Homo.sapiens)
class(Homo.sapiens)
Homo.sapiens
tx = transcripts(Homo.sapiens)
keytypes(Homo.sapiens)
columns(Homo.sapiens)
```

# Summary

We have covered a lot of material, from the nucleotide to the
pathway level.  The Annotation "view" at bioconductor.org
can always be visited to survey existing resources.
