# Gene set testing

In the following unit, we will explore software for testing differential expression in a set of genes. These tests differ from the gene-by-gene tests we saw previously. Again, the gene set testing software we will use lives in the `limma` package.

We download an experiment from the GEO website, using the `getGEO` function from the `GEOquery` package:

```{r}
library(GEOquery)
g <- getGEO("GSE34313")
e <- g[[1]]
```

This dataset is described at the following link:

http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE34313

Briefly, the investigators applied a glucocorticoid hormone to cultured human airway smooth muscle. The glucocorticoid hormone is used to treat asthma, as it reduces the inflammation response, however it has many other effects throughout the different tissues of the body.

The groups are defined in the `characteristics_ch1.2` variable:

```{r}
e$condition <- e$characteristics_ch1.2
levels(e$condition) <- c("dex24","dex4","control")
table(e$condition)
```

By examining boxplots, we can guess that the data has already been normalized somehow, and on the GEO site the investigators report that they normalized using Agilent software.

We will subset to the control samples and the samples treated with dexamethasone (the hormone) after 4 hours.

```{r}
boxplot(exprs(e), range=0)
names(fData(e))
lvls <- c("control", "dex4")
es <- e[,e$condition %in% lvls]
es$condition <- factor(es$condition, levels=lvls)
```

The following lines run the linear model in `limma`. We note that the top genes are common immune-response genes (CSF2, LIF, CCL2, IL6). Also present is FKBP5, a gene which regulates and is regulated by the protein which receives the glucocorticoid hormone.

```{r}
library(limma)
design <- model.matrix(~ es$condition)
fit <- lmFit(es, design=design)
fit <- eBayes(fit)
tt <- topTable(fit, coef=2, genelist=fData(es)$GENE_SYMBOL)
tt
```

We can test a single gene set by looking up the genes which contain a certain GO ID, and providing this to the `roast` function. We will show how to get such lists of genes associated with a GO ID in the next chunk.

The roast function performs an advanced statistical technique, *rotation of residuals*, in order to generate a sense of the null distribution for the test statistic. The test statistics in this case is the summary of the scores from each gene. The tests are *self-contained* because only the summary for a single set is used, whereas other gene set tests might compare a set to all the other genes in the dataset, e.g., a *competitive* gene set test.

The result here tells us that the *immune response* genes are significantly down-regulated, and additionally, mixed up and down.

```{r}
# Immune response
idx <- grep("GO:0006955", fData(es)$GO_ID)
length(idx)
r1 <- roast(es, idx, design)
?roast
r1
```

We can also use the `mroast` function to perform multiple roast tests. First we need to create a list, which contains the indices of genes in the ExpressionSet for each of a number of gene sets. We will use the `org.Hs.eg.db` package to gather the gene set information.

```{r}
# biocLite("org.Hs.eg.db")
library(org.Hs.eg.db)
org.Hs.egGO2EG
go2eg <- as.list(org.Hs.egGO2EG)
head(go2eg)
```

The following code unlists the list, then gets matches for each Entrez gene ID to the index in the ExpressionSet. Finally, we rebuild the list.

```{r}
govector <- unlist(go2eg)
golengths <- sapply(go2eg, length)
head(fData(es)$GENE)
idxvector <- match(govector, fData(es)$GENE)
table(is.na(idxvector))
idx <- split(idxvector, rep(names(go2eg), golengths))
go2eg[[1]]
fData(es)$GENE[idx[[1]]]
```

We need to clean this list such that there are no `NA` values. We also clean it to remove gene sets which have less than 10 genes.

```{r}
idxclean <- lapply(idx, function(x) x[!is.na(x)])
idxlengths <- sapply(idxclean, length)
idxsub <- idxclean[idxlengths > 10]
length(idxsub)
```

The following line of code runs the multiple ROAST test. This can take about 3 minutes.

```{r}
r2 <- mroast(es, idxsub, design)
head(r2)
r2 <- r2[order(r2$PValue.Mixed),]
```

We can use the `GO.db` annotation package to extract the GO terms for the top results, by the *mixed* test.

```{r}
# biocLite("GO.db")
library(GO.db)
columns(GO.db)
keytypes(GO.db)
GOTERM[[rownames(r2)[1]]]
r2tab <- select(GO.db, keys=rownames(r2)[1:10],
                columns=c("GOID","TERM","DEFINITION"), 
                keytype="GOID")
r2tab[,1:2]
```

We can also look for the top results using the standard p-value and in the *up* direction.

```{r}
r2 <- r2[order(r2$PValue),]
r2tab <- select(GO.db, keys=rownames(r2)[r2$Direction == "Up"][1:10],
                columns=c("GOID","TERM","DEFINITION"), 
                keytype="GOID")
r2tab[,1:2]
```

Again but for the *down* direction.

```{r}
r2tab <- select(GO.db, keys=rownames(r2)[r2$Direction == "Down"][1:5],
                columns=c("GOID","TERM","DEFINITION"), 
                keytype="GOID")
r2tab[,1:2]
```



