---
layout: page
title: Exploring Cufflinks output with cummeRbund
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```


Here we show the exploratory plots offered by the [cummeRbund](http://www.bioconductor.org/packages/release/bioc/html/cummeRbund.html) package. These plots require loading in a directory in which results from a [Cufflinks](http://cole-trapnell-lab.github.io/cufflinks/) analysis has been run. Follow the vignette in the above link in order in order to perform a Cufflinks gene- and isoform-level analysis. From the vignette:

> CummeRbund begins by re-organizing output files of a cuffdiff analysis, and storing these data in a local SQLite database. CummeRbund indexes the data to speed up access to specific feature data (genes, isoforms, TSS, CDS, etc.), and preserves the various relationships between these features. 

```{r message=FALSE}
library(cummeRbund)
myDir <- system.file("extdata", package="cummeRbund") 
gtfFile <- system.file("extdata/chr1_snippet.gtf",package="cummeRbund")
```

Read in the prepared Cufflinks files from the directory:

```{r warning=FALSE}
cuff <- readCufflinks(dir=myDir,gtfFile=gtfFile,genome="hg19",rebuild=TRUE)
```

Boxplots of expression (FPKM) at the gene and isoform level:

```{r}
csBoxplot(genes(cuff))
csBoxplot(genes(cuff),replicates=TRUE)
csBoxplot(isoforms(cuff),replicates=TRUE)
```

Scatterplot matrix of gene and isoform level expression:

```{r}
csScatterMatrix(genes(cuff))
csScatterMatrix(isoforms(cuff))
```

Sample dendrograms using Jensen-Shannon distances:

```{r}
csDendro(genes(cuff),replicates=TRUE)
csDendro(isoforms(cuff),replicates=TRUE)
```

MA-plot comparing two conditions:

```{r}
MAplot(genes(cuff),"hESC","Fibroblasts")
MAplot(isoforms(cuff),"hESC","Fibroblasts")
```

A "volcano plot" matrix. Each volcano plot is the -log10(p-value) over the log fold change.

```{r}
csVolcanoMatrix(genes(cuff))
csVolcanoMatrix(isoforms(cuff))
```

For all of these functions, see the help pages in the *cummeRbund* package for more details, and check the vignette for a sample workflow. The [Cufflinks homepage](http://cole-trapnell-lab.github.io/cufflinks/) has details about running the pipeline upstream of producing these figures.

```{r eval=FALSE}
browseVignettes("cummeRbund")
```
