# Homeworkd problems
Download the tables sample_tab.csv and tumor_tab.csv from https://github.com/genomicsclass/labs/tree/master/homeworks

The tables have clinical information from the same group of patients but each contains a different subset. Your task is to create a new talbe combining these tables for the samples that are in both

1. What is name of the column that contains a unique ID that can be used to map these two tables? Hint: it the name appears in both tables.

A) bcr_sample_uuid x
B) sample_type_id
C) bcr_sample_barcode

2. How many samples are represent in both tables

A) 51
B) 34 x
C) 37
D) 0

3. What is the average tumor weight of the samples that are i common?
A) 356.4118 x
B) 349.1622
C) NA

## Solutions

1. First we read in tables. We use as.is=TRUE because R turns characters into factors. We want characters.
```{r}
path <- "/Users/ririzarr/myDocuments/teaching/HarvardX/rawdata/homework"
tab1=read.csv(file.path(path,"sample_tab.csv"),as.is=TRUE)
tab2=read.csv(file.path(path,"tumor_tab.csv"),as.is=TRUE)
```
Now to find the column name that gives a unique ID first start by finding the column name the have in common

```{r}
intersect(names(tab1),names(tab2))
```
So there are only two possibilities. Looking at head of each we quicly see which one is unique.
```{r}
head(tab1$bcr_sample_uuid)
head(tab1$vial_number)
```
We can double check by checking uniqness
```{r}
any(duplicated(tab1$bcr_sample_uuid))
any(duplicated(tab2$bcr_sample_uuid))
```
So the answer is bcr_sample_uuid


2. We use the intersect function to find how many samples are in common. Note we can also use merge.
```{r}
common <- intersect(tab1$bcr_sample_uuid,tab2$bcr_sample_uuid)
length(common)
```

3. We first find which table contains tumor weight. We could look at all the column names but we will use grep to find the word "tumor"
```{r}
names(tab1)[grep("tumor",names(tab1))]
names(tab2)[grep("tumor",names(tab2))]
```
From here we see its the "tumor_weight" column in the second table. Now let's get the index of the sample that are in both. Then take the mean

```{r}
ind <- match(common,tab2$bcr_sample_uuid)
mean(tab2$tumor_weight)#[ind])
```

We wanted to show you how to use three very useful functions: interesct, match and grep but the quicker way to get the answer to 2 and 3 is the merge function
```{r}
newtab<-merge(tab1,tab2,by="bcr_sample_uuid")
nrow(newtab) ##number i common
mean(newtab$tumor_weight)
```




