---
layout: page
title: Principal Components Analysis
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## Principal Component Analysis 

We have already mentioned principal component analysis (PCA) above and noted its relation to the SVD. Here we provide further mathematical details. 

#### Example: Twin heights

We started the motivation for dimension reduction with a simulated example and showed a rotation that is very much related to PCA.


```{r simulate_twin_heights_again,echo=FALSE,message=FALSE,fig.cap="Twin heights scatter plot."}
library(rafalib)
library(MASS)

n <- 100

set.seed(1)

Y=t(mvrnorm(n,c(0,0), matrix(c(1,0.95,0.95,1),2,2)))

mypar()

thelim <- c(-3,3)
plot(Y[1,], Y[2,], xlab="Twin 1 (standardized height)", 
     ylab="Twin 2 (standardized height)", xlim=thelim, ylim=thelim)
points(Y[1,1:2], Y[2,1:2], col=2, pch=16)
```

Here we explain specifically what are the principal components (PCs).

Let $\mathbf{Y}$ be $2 \times N$ matrix representing our data. The analogy is that we measure expression from 2 genes and each column is a sample. Suppose we are given the task of finding a  $2 \times 1$ vector $\mathbf{u}_1$ such that $\mathbf{u}_1^\top \mathbf{v}_1 = 1$
and it maximizes $(\mathbf{u}_1^\top\mathbf{Y})^\top (\mathbf{u}_1^\top\mathbf{Y})$. This can be viewed as a projection of each sample or column of $\mathbf{Y}$ into the subspace spanned by $\mathbf{u}_1$. So we are looking for a transformation in which the coordinates show high variability.

Let's try $\mathbf{u}=(1,0)^\top$. This projection simply gives us the height of twin 1 shown in orange below. The sum of squares is shown in the title.

```{r projection_not_PC1, fig.align="Data projected onto space spanned by (1 0).", fig.width=6.5,fig.height=6.5}
mypar(1,1)
plot(t(Y), xlim=thelim, ylim=thelim,
     main=paste("Sum of squares :",round(crossprod(Y[1,]),1)))
abline(h=0)
apply(Y,2,function(y) segments(y[1],0,y[1],y[2],lty=2))
points(Y[1,],rep(0,ncol(Y)),col=2,pch=16,cex=0.75)
```

Can we find a direction with higher variability? How about:

$\mathbf{u} =\begin{pmatrix}1\\-1\end{pmatrix}$ ? This does not satisfy $\mathbf{u}^\top\mathbf{u}= 1$ so let's instead try
$\mathbf{u} =\begin{pmatrix}1/\sqrt{2}\\-1/\sqrt{2}\end{pmatrix}$ 

```{r projection_not_PC1_either, fig.cap="Data projected onto space spanned by (1 0).", fig.width=6.5,fig.height=6.5}
u <- matrix(c(1,-1)/sqrt(2),ncol=1)
w=t(u)%*%Y
mypar(1,1)
plot(t(Y),
     main=paste("Sum of squares:",round(tcrossprod(w),1)),xlim=thelim,ylim=thelim)
abline(h=0,lty=2)
abline(v=0,lty=2)
abline(0,-1,col=2)
Z = u%*%w
for(i in seq(along=w))
  segments(Z[1,i],Z[2,i],Y[1,i],Y[2,i],lty=2)
points(t(Z), col=2, pch=16, cex=0.5)
```

This relates to the difference between twins, which we know is small. The sum of squares confirms this.

Finally, let's try:

$\mathbf{u} =\begin{pmatrix}1/\sqrt{2}\\1/\sqrt{2}\end{pmatrix}$ 

```{r PC1, fig.cap="Data projected onto space spanned by first PC.", fig.width=6.5,fig.height=6.5}
u <- matrix(c(1,1)/sqrt(2),ncol=1)
w=t(u)%*%Y
mypar()
plot(t(Y), main=paste("Sum of squares:",round(tcrossprod(w),1)),
     xlim=thelim, ylim=thelim)
abline(h=0,lty=2)
abline(v=0,lty=2)
abline(0,1,col=2)
points(u%*%w, col=2, pch=16, cex=1)
Z = u%*%w
for(i in seq(along=w))
  segments(Z[1,i], Z[2,i], Y[1,i], Y[2,i], lty=2)
points(t(Z),col=2,pch=16,cex=0.5)
```

This is a re-scaled average height, which has higher sum of squares. There is a mathematical procedure for determining which $\mathbf{v}$ maximizes the sum of squares and the SVD provides it for us.

#### The principal components

The orthogonal vector that maximizes the sum of squares:

$$(\mathbf{u}_1^\top\mathbf{Y})^\top(\mathbf{u}_1^\top\mathbf{Y})$$ 

$\mathbf{u}_1^\top\mathbf{Y}$ is referred to as the first PC. The _weights_ $\mathbf{u}$ used to obtain this PC are referred to as the _loadings_. Using  the language of rotations, it is also referred to as the _direction_ of the first PC, which are the new coordinates.

To obtain the second PC, we repeat the exercise above, but for the residuals:

$$\mathbf{r} = \mathbf{Y} - \mathbf{u}_1^\top \mathbf{Yv}_1 $$

The second PC is the vector with the following properties: 

$$ \mathbf{v}_2^\top \mathbf{v}_2=1$$

$$ \mathbf{v}_2^\top \mathbf{v}_1=0$$ 

and maximizes  $(\mathbf{rv}_2)^\top \mathbf{rv}_2$.

When $Y$ is $N \times m$ we repeat to find 3rd, 4th, ..., m-th PCs.

#### `prcomp`

We have shown how to obtain PCs using the SVD. However, R has a function specifically designed to find the principal components. In this case, the data is centered by default. The following function: 

```{r}
pc <- prcomp( t(Y) )
```

produces the same results as the SVD up to arbitrary sign flips:

```{r pca_svd, fig.cap="Plot showing SVD and prcomp give same results.",fig.width=10.5,fig.height=5.25}
s <- svd( Y - rowMeans(Y) )
mypar(1,2)
for(i in 1:nrow(Y) ){
  plot(pc$x[,i], s$d[i]*s$v[,i])
}
```

The loadings can be found this way:
```{r}
pc$rotation
```
which are equivalent (up to a sign flip) to:
```{r}
s$u
```
The equivalent of the variance explained is included in the: 
```{r}
pc$sdev
```
component.



We take the transpose of `Y` because `prcomp` assumes the previously discussed ordering: units/samples in row and features in columns.




