---
layout: page
title: "A view of genetic heterogeneity between and within cancer types"
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
library(ph525x)
library(RTCGAToolbox)
})
```


## Introduction

We will use data in the ph525x package on mutations in
breast cancer and rectal adenocarcinoma to illustrate
some issues in dealing with mutations data from TCGA.
A basic objective is construction of a "rainfall plot".
An example is Figure 6 from [Alexandrov et al. 2013](http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3776390&tool=pmcentrez&rendertype=abstract):

```{r lkkat,fig=TRUE,echo=FALSE}
kataegis()
```

These plots include data from deeply sequenced individual tumors,
and we'd like to understand how to construct them using
tools from Bioconductor.

## The mutation data frames from RTCGAToolbox

The `readMuts` data are from the 20150402 TCGA production.
```{r lkread}
library(ph525x)
data(readMuts)
dim(readMuts)
data(brcaMuts)
dim(brcaMuts)
```

## Mutation types and their contents

```{r lkrmut}
table(readMuts$Variant_Type)
with(readMuts, head(Reference_Allele[Variant_Type=="DEL"]))
```

## Tabulating substitution types

The following function enumerates substitutions according to
the [COSMIC convention](http://cancer.sanger.ac.uk/cosmic/signatures):
"The profile of each signature is displayed using the six substitution subtypes: C>A, C>G, C>T, T>A, T>C, and T>G (all substitutions are referred to by the pyrimidine of the mutated Watson–Crick base pair)."

```{r dosubt}
 subt = function(ref, a1, a2) {
        alt = ifelse(a1 != ref, a1, a2)
        tmp = ref
        needsw = which(alt %in% c("C", "T"))
        ref[needsw] = alt[needsw]
        alt[needsw] = tmp[needsw]
        paste(ref, alt, sep = ">")
    }
with(readMuts[readMuts$Variant_Type=="SNP",],
   table(subt(Reference_Allele, Tumor_Seq_Allele1, Tumor_Seq_Allele2)))
```

A>G and G>A substitutions are not included in kataegis plots.

To define the colors used for substitutions:

```{r lkkac}
ph525x:::kataColors
```

## Total genomic distance

The mutation locations reported are not particularly convenient for genome-wide
plotting as the distances are all relative to chromosome start.
The following hidden function computes total distance relative
to start of chr1, assuming that the data are held in GRanges.
```{r lktg}
ph525x:::totalgd
```

## A demo plot for four tumors

The rainfall function will organize the input data by sample, and
samples can, in the present version, be selected according to
their position in an ordering based on number of mutations reported.
The default plots the sample with the greatest number of mutations.
The oind parameter allows selection of samples further down in the
ordering.  We embellish the plot with a simple kernel estimate
of the density of mutations along the chromosomes.  The
function invisibly returns a list of items related to the plot.

```{r do4f,fig=TRUE,fig.height=8}
rainouts = list()
par(mfrow=c(4,1),mar=c(4,5,1,1))
for (i in 1:4) rainouts[[i]] = rainfall(readMuts, oind=i)
```

```{r lkrao}
str(rainouts[[1]])
```


