---
layout: page
title: Bioconductor for genome-scale data -- quick intro
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r setup,echo=FALSE,results="hide"}
suppressWarnings({
suppressPackageStartupMessages({
library(Biobase)
library(GSE5859)
library(annotate)
library(BiocParallel)
library(VariantAnnotation)
library(BSgenome.Hsapiens.UCSC.hg19)
})
})
```

# PH525.6x: Basic premise and overview

You know to manipulate and analyze data using R, and
you understand a considerable amount about statistical modeling.
If you've taken PH525.5x, you've gotten significant background
on current agendas in computational biology, and have learned
how to deal with genomic data from the management, annotation,
and analysis perspectives.

In this course, we will use Bioconductor as the foundation
of demonstrations and exercises in
* methods for interactive visualization of genome-scale data with the shiny and ggvis packages;
* programming strategies for scalable bioinformatics with multicore and cluster computing infrastructure;
* integrative management and analysis of multiassay experiments, with illustrations from The Cancer Genome Atlas (TCGA);
* approaches to improving reproducibility of genome-scale analyses;

One week will be devoted to each of these topics.
