---
layout: page
title: "Combining S4 with NoSQL (mongodb) to query ENCODE bedfiles"
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
library(Biobase)
library(TxRegInfra)
library(ph525x)
library(BiocStyle)
})
```


## Introduction

For an application involving many thousands of files,
we have found that [NoSQL](https://en.wikipedia.org/wiki/NoSQL)
strategies may be effective.  The [TxRegInfra](https://github.com/vjcitn/TxRegInfra) package is under development in github 
and illustrates use of mongodb
with a small collection of BED files obtained from the
ENCODE project.  In this section we sketch the most
basic aspects of wrapping a mongodb connection in S4 and
implementing subsetByOverlaps to query the data store.

To carry out the tasks in this section, you will need mongod
(the database managing daemon) running on your system.  
The [community server edition](https://www.mongodb.com/download-center?jmp=nav#community) should be easy to install.

After you get mongod running, you can install TxRegInfra using
`library(BiocInstaller); biocLite("vjcitn/TxRegInfra")`.

## Basic considerations

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(TxRegInfra)
library(GenomicFiles)
})
```

Our long term goal is
to define a package,
TxRegQuery, to exploration of transcriptional regulatory networks
by integrating data on eQTL, digital genomic footprinting (DGF), DnaseI
hypersensitivity binding data (DHS), and transcription
factor binding site (TFBS) data.  Owing to the volume of emerging tissue-specific
data, special data modalities are used.  In this document
we'll focus on DHS.

## Managing bed file content with mongodb

### Importing and querying documents

The package comes with a small number of bed files to demonstrate
import utilities.
```{r lkk}
# ENCODE
f1 = dir(system.file("bedfiles", package="TxRegInfra"), full=TRUE, patt="ENCFF971VCD")
cat(readLines(f1, n=3), sep="\n")
# ChromHMM
f2 = dir(system.file("bedfiles", package="TxRegInfra"), full=TRUE, patt="E096_imp12")
cat(readLines(f2, n=3), sep="\n")
```

The function `importBedToMongo` uses system() to run
mongodb.
There is a `bedType` parameter that indicates what fields are available; it
defaults to `broadPeak`.

The following code imports a broadPeak and chromHMM document.
We deal with metadata about these documents below.
We assume a database called 'txregnet' has been established
for a running mongodb server.
```{r doimp}
importBedToMongo(f1, "vjc1", db="txregnet")
importBedToMongo(f2, "vjc2", db="txregnet", bedType="chromHMM")
```

Now that the documents are imported, we can query for
information in an interval specified by a GRanges instance.
```{r doq}
library(RMongo)
con = mongoDbConnect("txregnet") # defaults for local server
queryBedInMongo(con, "vjc1", GRanges("chr1", IRanges(1, 800000)), skip=0, limit=5)
queryBedInMongo(con, "vjc2", GRanges("chr17", IRanges(1, 800000)), skip=0, limit=5)
```

## An integrative container

We need to bind the metadata and information about the mongodb.

### BED file metadata

The BED files are extracted from a few different places.  We have
metadata on 10 of them:
```{r lkmeta}
data(hsFiles_subset) # holds hsFiles
hsFiles[1:3,1:6]
```
We added an additional four.  This will become colData for an
instance of an extended RaggedExperiment class to be defined.
```{r donew,echo=FALSE}
library(S4Vectors)
e072 = data.frame(File.accession = "E072_imp12_mn_trun",
   File.format = "bed ChromHMM", Output.type = "states", Experiment.accession=NA,
    Assay = "ChromHMM", Biosample.term.id=NA, 
    Biosample.term.name="brain inf. temporal lobe",
    Biosample.type=NA,
    Biosample.life.stage=NA, Biosample.sex=NA)
e073 = data.frame(File.accession = "E073_imp12_mn_trun",
   File.format = "bed ChromHMM", Output.type = "states", Experiment.accession=NA,
    Assay = "ChromHMM", Biosample.term.id=NA, 
    Biosample.term.name="brain prefr. cortex",
    Biosample.type=NA,
    Biosample.life.stage=NA, Biosample.sex=NA)
e088 = data.frame(File.accession = "E088_imp12_mn_trun",
   File.format = "bed ChromHMM", Output.type = "states", Experiment.accession=NA,
    Assay = "ChromHMM", Biosample.term.id=NA, 
    Biosample.term.name="fetal lung",
    Biosample.type=NA,
    Biosample.life.stage=NA, Biosample.sex=NA)
e096 = data.frame(File.accession = "E096_imp12_mn_trun",
   File.format = "bed ChromHMM", Output.type = "states", Experiment.accession=NA,
    Assay = "ChromHMM", Biosample.term.id=NA, 
    Biosample.term.name="adult lung",
    Biosample.type=NA,
    Biosample.life.stage=NA, Biosample.sex=NA)
```
```{r lkcd}
cd = DataFrame(rbind(hsFiles, rbind(e072, e073, e088, e096)))
cd[1:4,1:6]
```

### S4: Extending the RaggedExperiment class

(From the RaggedExperiment vignette:)
The `r Biocpkg("RaggedExperiment")` package provides a flexible data
representation for copy number, mutation and other ragged array schema for 
genomic location data. It aims to provide a framework for a set of samples
that have differing numbers of genomic ranges.

In TxRegInfra, we extend the RaggedExperiment class to deal
with external data managed by mongodb.  We've created
a database 'txregnet' and we connect this to 
the extended RaggedExperiment
'rme1', an instance of `RaggedMongoExpt`.

```{r domor1}
okdf = DataFrame(hsFiles)
rownames(okdf) = hsFiles[,1]
loccon = localMongolite(db="txregnet")
rme1 = RaggedMongoExpt(loccon, colData=okdf)
rme1
```

## The upshot: peak densities by tissue type

In the following, we produce a table of number of
peaks by tissue type, in a small region of chromosome 1.

```{r ccc}
brp = which(colData(rme1)$File.format == "bed broadPeak")
allst = subsetByOverlaps(rme1[,brp], 
               GRanges("chr1", IRanges(1,8e5))) 
data.frame(tiss=colData(rme1)[brp, "Biosample.term.name"], 
             num.peaks=sapply(allst,nrow))
```

## Some additional details

Ultimately we would like to make use of the RaggedExperiment
infrastructure directly.  To do this we need to bind
a GRangesList to the assay data; once this is done,
we can use the sparseAssay, compactAssay, and qreduceAssay
methods.  Longer term utility of this approach will be
demonstrated in the TxRegQuery package, under development.

```{r tst}
badn = c("seqnames", "ranges", "strand", "seqlevels", 
   "seqlengths", "isCircular", "start", "end", "width", "element")
cleanCols = function(x) setdiff(colnames(x), badn)
grl = GRangesList(lapply(allst, function(x) {
     ans = GRanges(x$chrom, IRanges(x$chromStart, x$chromEnd)); mcols(ans) = x[,cleanCols(x)]; ans
     }))
re = RaggedExperiment(grl, colData=colData(rme1[,brp])) 
dim(sparseAssay(re))
dim(compactAssay(re))
```

To conclude, we peek at the details of the mongodb connection
established by `r CRANpkg("mongolite")`.  It includes a
variety of hints concerning the R interface.

```{r dddd}
rme1@con
rme1@con@con
```
