---
layout: page
title: shiny for interactive visualization of multivariate data
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r setup,echo=FALSE,results="hide"}
suppressWarnings({
suppressPackageStartupMessages({
library(shiny)
library(cluster)
library(ph525x)
})
})
```

## Running the dfHclust function

### The main display layout

The ph525x package includes the `dfHclust` function.  This
takes a `data.frame` instance as sole argument and starts
a browser session.  You can try it directly with

```{r lkb,eval=FALSE}
library(ph525x)
dfHclust(mtcars)
```

There is a sidebar panel on the left that accepts selections
for

* `distance` for object proximity
* `clustering method` to form the hierarchy through object agglomeration
* `height` for cutting the cluster dendrogram into groups of objects
* features to use for computation of object distance

In this application, the objects are cars with different manufacturer
and model types; the features are structural or operating characteristics
of cars.  Note that the application starts with certain defaults:

* euclidean distance
* ward.D as the agglomeration procedure
* 40 as the cut height
* the first two variables as features

The main panel to the right includes tabs for three types of display:

* tree for the clustering dendrogram
* pairs plot for direct pairwise scatterplot visualization 
* silhouette plot for assessing cluster quality at the selected cutting height

To fully understand the quantities displayed in the "tree" and "silh"
tabs, you should review the definitions of distance, clustering method, 
and silhouette until you feel comfortable explaining these to a non-statistician.

* Distance measures are very important in multivariate analysis
* Hierarchical clustering procedures are complex; see the 
[definition of Ward's method](https://en.wikipedia.org/wiki/Ward%27s_method) in Wikipedia for a clear exposition of one important approach
* Silhouette is defined clearly in the man page for `silhouette` in the `cluster` package

### Interacting with the display

Notice that the defaults lead to a tree with three main lobes in the
panel displayed when the 
"tree" tab is selected.  The panel for the "silh" tab shows 
a measure of cluster membership for each observation in the dataset.

* note that the average silhouette width at the defaults for mtcars is 0.56
* if, leaving all other selections alone, you change the "Select height for cut" value to 70, the silhouette plot changes to show an average silhouette of 0.59 -- an improvement, but now we have only two clusters, and one observation with a negative silhouette value
* for an amusing observation, set the distance to "euclidean" and the clustering method to `single` -- now the closest neighbor to the Mercedes 450 SLC is the AMC Javelin.  I guess the Javelin wasn't such a bad car after all....

