---
layout: page
title: shiny for interactive visualization of multivariate data
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r setup,echo=FALSE,results="hide"}
suppressWarnings({
suppressMessages({
suppressPackageStartupMessages({
library(shiny)
library(cluster)
library(ph525x)
})
})
})
```

## Interaction beyond the "read-eval-print" loop

Working do with R through the command line imposes
a certain laborious sequentiality to everything we do.
The advantage to this mode of data-analytic computing is
that every step is defined by code and every result can be
viewed (and traced) as a sequence of explicit function evaluations on
a well-defined environment.

An alternative mode of data-analytic computing uses
graphical user interfaces (GUIs) in which mouse movements
and button clicks are the primary means of specifying 
computations.  This mode of interaction is sequential and
traceable as well, but the path from mouse click to data-analytic
function is technically complex and we will not pursue the
matter any further.

The Rstudio group has produced a package called `shiny` [URL](shiny.rstudio.com) that
simplifies the creation of browser-driven GUIs for specific
data-analytic activities.  In this lab we'll explore some of
the potential of this package.  We've added a "shiny app" to
the ph525x package that we'll now investigate.

## Running the dfHclust function

### The main display layout

The ph525x package includes the `dfHclust` function.  This
takes a `data.frame` instance as sole argument and starts
a browser session.  You can try it directly with

```{r lkb,eval=FALSE}
library(ph525x)
dfHclust(mtcars)
```

There is a sidebar panel on the left that accepts selections
for

* `distance` for object proximity
* `clustering method` to form the hierarchy through object agglomeration
* `height` for cutting the cluster dendrogram into groups of objects
* features to use for computation of object distance

In this application, the objects are cars with different manufacturer
and model types; the features are structural or operating characteristics
of cars.  Note that the application starts with certain defaults:

* euclidean distance
* ward.D as the agglomeration procedure
* 40 as the cut height
* the first two variables as features

The main panel to the right includes tabs for three types of display:

* tree for the clustering dendrogram
* pairs plot for direct pairwise scatterplot visualization 
* silhouette plot for assessing cluster quality at the selected cutting height

To fully understand the quantities displayed in the "tree" and "silh"
tabs, you should review the definitions of distance, clustering method, 
and silhouette until you feel comfortable explaining these to a non-statistician.

* Distance measures are very important in multivariate analysis
* Hierarchical clustering procedures are complex; see the 
[definition of Ward's method](https://en.wikipedia.org/wiki/Ward%27s_method) in Wikipedia for a clear exposition of one important approach
* Silhouette is defined clearly in the man page for `silhouette` in the `cluster` package

### Interacting with the display

Notice that the defaults lead to a tree with three main lobes in the
panel displayed when the 
"tree" tab is selected.  The panel for the "silh" tab shows 
a measure of cluster membership for each observation in the dataset,
using the height-to-cut setting to partition the data using the tree.

* note that the average silhouette width at the defaults for mtcars is 0.56
* if, leaving all other selections alone, you change the "Select height for cut" value to 70, the silhouette plot changes to show an average silhouette of 0.59 -- an improvement, but now we have only two clusters, and one observation with a negative silhouette value
* for an amusing observation, set the distance to "euclidean" and the clustering method to `single` -- now the closest neighbor to the Mercedes 450 SLC is the AMC Javelin.  I guess the Javelin wasn't such a bad car after all....

## A view of the code

In this subsection we will break up the main `dfHclust` function and
explain its elements.  The code was written in a very naive manner but
even so has three virtues:

* it is relatively short and self-contained
* it works and does something useful in interactive data exploration
* it is easy to extend by replicating and modifying short subparts

Some aspects that are likely to need improvement

* handling of global variables for use in the `ui` component
* excessive repetition of data.frame subsetting in the `server` component

### Starting out

We have a simple fixed interface and fail if we don't get a data.frame
with at least two columns.  We fail if the necessary software is
not in place.
```
dfHclust = function(df) {
# validate input
   stopifnot(inherits(df, "data.frame"))
   stopifnot(ncol(df)>1)
# obtain software
   require(shiny)
   require(cluster)
```
### Some vectors to populate interface option sets

```
# global variables ... 
   nms = names(df)
   cmeths = c("ward.D", "ward.D2",
             "single", "complete", "average", "mcquitty",
             "median", "centroid")
   dmeths = c("euclidean", "maximum", "manhattan", "canberra",
             "binary")
```

### The layout of the user interface

There are several high-level options that can control
the browser interface.  We'll use a flexible approach
called `fluidPage`; see also the shinydashboard package.

```
#
# main shiny components: ui and server
#   ui: defines page layout and components
#   server: defines operations
#
   ui <- fluidPage(
#
# we will have four components on sidebar: selectors for 
# distance, agglomeration method, height for tree cut, and variables to use
```

#### Title and sidebar
```
#
     titlePanel(paste(substitute(df), "hclust")),
     sidebarPanel(
          helpText(paste("Select distance:" )),
          fluidRow(
             selectInput("dmeth", NULL, choices=dmeths,
               selected=dmeths[1])),
          helpText(paste("Select clustering method:" )),
          fluidRow(
             selectInput("meth", NULL, choices=cmeths,
               selected=cmeths[1])),
          helpText(paste("Select height for cut:" )),
          fluidRow(
             numericInput("cutval", NULL, value=40, min=0, max=Inf, step=1)),
          helpText(paste("Select variables for clustering from", substitute(df), ":" )),
          fluidRow(
             checkboxGroupInput("vars", NULL, choices=nms,
               selected=nms[1:2]))
            ),
```
#### Tabs for the primary displays

```
#
# main panel is a simple plot
#
     mainPanel(
       tabsetPanel(
        tabPanel("tree", 
         plotOutput("plot1")),
        tabPanel("pairs", 
         plotOutput("pairsplot")),
        tabPanel("silh", 
         plotOutput("silplot"))
         )
       )
  )  # end fluidPage
```

### Defining the server
   
```
#
# a function with up to three arguments (input, output, session)
# can be used to define the server component of the app
# renderPlot makes it reactive, so when input components are altered,
# data frame in use and plot are updated
#
   server <- function(input, output) {
     output$plot1 <- renderPlot({
       xv = df[,input$vars]
       plot(hclust(dist(data.matrix(xv),method=input$dmeth), method=input$meth),
         xlab=paste(input$dmeth, "distance;", input$meth, "clustering"))
       abline(h=input$cutval, lty=2, col="gray")
     })
     output$pairsplot <- renderPlot({
       xv = df[,input$vars]
       pairs(data.matrix(xv))
     })
     output$silplot <- renderPlot({
       xv = df[,input$vars]
       dm = dist(data.matrix(xv),method=input$dmeth)
       hc = hclust(dist(data.matrix(xv),method=input$dmeth), method=input$meth)
       ct = cutree(hc, h=input$cutval)
       plot(silhouette(ct, dm))
     })
   }
   
   shinyApp(ui, server)
}
```

## Some gory details

A "shiny app" consists of two main components, a user interface and
a server function.  To use the infrastructure the `shiny`
library must be attached to an R session.
The application can be started in various ways; we use
`shinyApp` and supply two arguments, `ui` and `server`.

* `ui` is an instance of `shiny.tag.list`.  You can get a feel
for this by inspecting the result of `fluidPage()`.

* `server` is a function of three arguments `input`, `output`
and `session`; the latter is optional.  `input` is a list
with bindings given values in the `ui` component, and `output`
will be populated with elements in the server for rendering in
the UI.

### An example with fluidPage

To appreciate what shiny is doing for you, try the following:

```{r liks, eval=FALSE}
library(shiny)
fluidPage(
  titlePanel("a title"),
  sidebarPanel(
    selectInput("seli1", "some letters", letters[1:4])
    )
  )
```

Set `sink(file="dem.html")` and run the above code again.  Then
issue `sink(NULL); browseURL("dem.html")`.  R will fire up the browser
and show a fairly anemic page.  There will be a select control with
options a, ..., d.  This particular example would be easy to code by
hand, but the `selectInput` function allows you to develop controls
with choices defined by any R vector.  Furthermore, high-level
R functions are defined to allow different kinds of input, which
may be driven by mouse events.
