---
layout: page
title: "Sharded GRanges: a hybrid in/out of memory strategy for large sets of ranges"
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
library(Biobase)
library(geuvStore)
library(gQTLBase)
library(gQTLstats)
library(foreach)
library(doParallel)
library(ph525x)
})
```


## Introduction

We've looked at a number of approaches to
working with data external to R:

* HDF5, which manages groups of multidimensional arrays on disk
* sqlite, a zero-configuration relational database
* tabix, a simple approach to indexing records on genomic coordinates

Here I want to describe an approach that seems useful for millions
of ranges annotated in the course of searching for variants that
affect gene expression at the population level.  The approach
is based on a concept of storing data in "shards", homogeneous small
fragments that can be quickly loaded and unloaded, discoverable
by index and traversable in parallel.

## Motivation: An integrative view of associations in GEUVADIS

The [GEUVADIS study](http://www.nature.com/nature/journal/v501/n7468/full/nature12531.html) is an intensive multiomic study of gene expression in multiple
populations.  We want to make use of the data from this study to
investigate variants affecting genes of interest, with one tool
an interactive graphical utility illustrated in the video:

```{r lkgg,fig=TRUE}
library(ph525x)
ggshot()
```

We want to be able to select genes by symbol and explore names
and epigenetic contexts of variants whose content is associated with
expression variation.  It is useful to have the variants annotated
using GRanges, but a very large GRanges object (there are hundreds
of millions of SNP-gene associations recorded) can be unwieldy.
Solutions using RDBMS or HDF5 may be viable but more infrastructure
for rapidly searching such stores using genomic coordinates,
and for converting query results to GRanges will be needed.

BatchJobs was used to generate the association tests, and it
produces
an organized system of "sharded" GRanges recording the
associations along with metadata about the associated features.
This system can be stored in a package, exemplified by geuvStore.

## A quick look at geuvStore

The association test results are organized using a BatchJobs
registry that is wrapped in an S4 class called ciseStore.
```{r lkgv}
library(geuvStore)
m = makeGeuvStore()
class(m)
m
```

The show method for m probes into the store and retrieves one record
from one GRanges instance.

## Scalable traversal

The traversal of all GRanges available in this selection is
governed by foreach loops.
```{r lksca, cache=TRUE}
library(gQTLBase)
ut1 = unix.time(l1 <- storeApply(m, length))
ut1
library(doParallel)
registerDoParallel(cores=2)
ut2 = unix.time(l2 <- storeApply(m, length))
ut2
print(sum(unlist(l2)))
all.equal(unlist(l1), unlist(l2))
```
We see that doubling the number of processors halves the
time required to get the length of each component of the archive.
With large numbers of cores, we can quickly assemble information
about many variants.

## Scalable histogram construction

When the histogram bins are fixed, divide and conquer can be
used to assemble a histogram in parallel over many chunks.

```{r lkhhh,cache=TRUE}
registerDoParallel(cores=1)
unix.time(ll <- storeToHist(m, getter=function(x)log(mcols(x)$chisq+1), breaks=c(0,seq(.1,5,.1),10)))
registerDoParallel(cores=2)
unix.time(ll <- storeToHist(m, getter=function(x)log(mcols(x)$chisq+1), breaks=c(0,seq(.1,5,.1),10)))
```

## Indexing for targeted retrievals

The ciseStore class includes two maps: one from range to shard number,
another from gene identifier to shard number.  This allows rapid
retrievals.

```{r lkex}
myr = GRanges(2, IRanges(1975.7e5, width=50000))
extractByRanges(m, myr)
```

## Conclusions

geuvStore is a complex architecture that aims to provide a
partly baked representation of quantities from genome-scale
surveys that can be scalably surveyed and integrated.  This
is accomplished by keeping ranges for association scores
and metadata in small sharded GRanges with some simple indexes,
retrieval utilities, and with support for parallelized traversal
and summary.  It would be very nice to achieve these aims with
a more homogeneous underlying architecture such as HDF5, and
this may be possible as file-backed SummarizedExperiments come
on line.

