---
layout: page
title: "Some final comments on genome-scale visualization"
---

```{r options, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
suppressWarnings({
library(BiocStyle)
library(Biobase)
library(randomForest)
library(MLInterfaces)
library(tissuesGeneExpression)
library(limma)
library(ph525x)
library(RCircos)
})
})
```


## RCircos

RCircos is not distributed in Bioconductor, but can
be useful for developing compact displays of interactions
among genomic elements.  I am unaware of any interfaces between
Bioconductor data classes and RCircos, and this topic deserves
attention.

In the ph525x package we have added a selection of trans-eQTL
findings from Westra et al. Nature 2013 (doi: 10.1038/ng.2756).
We show a few SNP-gene associations from this study:
```{r lksn,fig=TRUE,message=FALSE,fig.height=9,fig.width=9}
library(ph525x)
data(westraTransSel)
westraTransSel[1:3]
sglToCircos(westraTransSel[1:5])
```

## ComplexHeatmap

`r Biocpkg("ComplexHeatmap")` has a very nice vignette addressing many
issues in combining heatmaps and repurposing the heatmap 
concept.  The oncoprint example in the vignette is particularly
comrelling.  To use this interactively with TCGA, contact
[the ISB](http://isb-cancer-genomics-cloud.readthedocs.io/en/latest/sections/FAQ.html) and obtain a cloud platform account.
Then obtain the `r Biocpkg("cgcR")`
package, load it, and run `isbApp()`.  You will have to authenticate
with google to get access to the BigQuery representation of TCGA.

## WebGL and interaction with data

In the short concluding video we use the MLInterfaces plspinHcube
function to illustrate several aspects of interactivity: GUI for
tuning, mouse-controlled rotation, and mouseover for point interrogation.

## EpiViz

The `r Biocpkg("epivizr")` package interacts with the
[epiviz](https://epiviz.github.io/) system and is capable of substantial feats of data integration and
higher-level data interactivity.
