---
layout: page
title: "Benchmarking multiple out-of-memory strategies"
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
suppressPackageStartupMessages({
suppressWarnings({
library(ph525x)
library(microbenchmark)
library(bigmemory)
library(rhdf5)
library(RSQLite)
})
})
```


## Introduction

In many large-data situations, it is impractical to load and retain
data in R's working memory space.  We have had a look at
HDF5, SQLite and tabix-indexed text as possible solutions
to problems arising with memory constraints.    We'll call
these "out-of-memory" (OOM) approaches

How can we obtain data on which approach will be most effective
for a given task?  Comparative benchmarking is a very useful skill and
we give a very rudimentary account of this here.

## The harness

It is common to speak of a program that drives other programs
as a "harness" (see [wikipedia](https://en.wikipedia.org/wiki/Test_harness)
for related discussion).  We have such a program in ph525x:

```{r lkph}
benchOOM
```

This program is going to help us assess performance of various
OOM approaches.  We consider a very limited problem, that of
managing data that could reside in an R matrix.
The main parameters are

* `NR` and `NC`: row and column dimensions
* `times`: number of benchmark replications for averaging
* `inseed`: a seed for random number generation to ensure reproducibility
* `methods`: a list of methods

The `methods` parameter is most complex.  Each element of the list
is assumed to be a function with the matrix to
be managed via OOM as the first argument, some additional
parameters, and a parameter `intimes` that gives the number 
of benchmark replicates.

Our objective is to produce a table that looks like

```
> b1
    NR  NC times       meth        wr     ingFull      ing1K
1 5000 100     5       hdf5  10.71714   9.4100810 14.2984402
2 5000 100     5         ff  25.34365  63.0977338  4.4320688
3 5000 100     5     sqlite 174.89003 105.1254638 28.4717496
4 5000 100     5 data.table  49.35190   7.9871552 13.9007588
5 5000 100     5  bigmemory  23.39697   0.9660878  0.9950034
```

where each method listed in `meth` is asked to perform the same
task a fixed number of times for averaging.  The construction of 
the table occurs by binding together metadata about the task and
method to the result of `getStats`.  We'll leave the details
of `getStats` to independent investigation.


## An example method (OOM benchmarker)

Let's look at the method for HDF5:
```{r lkme}
ph525x:::.h5RoundTrip
```

The program has three main phases

* HDF5-related setup, cleaning out any previous archives and establishing
the basic target file
* Benchmarking of data export via `h5write`
* Benchmarking of ingestion via `h5read` with various restrictions

The results of `microbenchmark` are assembled in a list.
