---
layout: page
title: Introduction to random variables and reading data
---

# Introduction

```{r,results=FALSE,echo=FALSE}
set.seed(1) ##so that we get same results
```

This course introduces the statistical concepts necessary to underatadn p-value and confidence intervals that are ubiquoitous in the life science literature. Let's take an example. This paper [http://diabetes.diabetesjournals.org/content/53/suppl_3/S215.full] states that 

"Body weight was higher in mice fed the high-fat diet already after the first week, due to higher dietary intake in combination with lower metabolic efficiency. "

To back this up they provide this

"Already during the first week after introduction of high-fat diet, body weight increased significantly more in the high-fat diet–fed mice (+1.6 ± 0.1 g) than in the normal diet–fed mice (+0.2 $\pm$ 0.1 g; P < 0.001)."

What does P < 0.001 mean? What are $\pm$ included? In this class we will learn what this mean and learn to compute these values in R. The first step is to understand what is a random variable. To understand this, we will use data from a mouse database [CITE: provided by Karen Svenson via Gary Churchill and <Gary.Churchill@jax.org> Dan Gatti and Partially funded by P50 GM070683.]. We will import the data with R and explain random variables and null distributions using R programming.

## Importing data into R
Before defining random variable and null distribution let's import data into R.

In the life sciences, small datasets such as the one used as an example here are stored as Excel file. In general you want to avoid the Excel (xls) format and save files as comma delimited (csv) or tab delimted (txt) files. The first step is to find this file and know it's *path*. 

### Finding the file

The file we are using can be found our github repository [here](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/miceweights.csv). Now once you download the file, how do you find the path? Windows and Macs have slighlty different ways of finding these. I'll leave to you to figure this out. 

You can also use R to downloade the file and define the path yourself. Let's download the file and put it in our Desktop. In general we don't recommend downloading files to your Desktop. We use it as an example.

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
```

Now we are ready to read in the file 
```{r}
dat <- read.csv(filename)
```

## Our first look at data

We are interesting determing if following a given diet makes mice heavier after several. This data was produced by ordering 24 mice from Jackson Lab, randomely assigning either chow or high fat (hf) diet. Then after several weeks we weighed each mice and obtained this data:

```{r}
dat
```

So are the hf mice heavier? Note that mouse 21 is 20.73 grams is one the lightest mice while 14 at 34.02 is one of the heaviest. Both are on the hf diet. Just from looking the data we see there is *variability*. Claims such as the one above usually refer to the averages. So let's look at the average of each grou

```{r}
control <- dat[1:12,2]
treatment <- dat[13:14,2]
print(mean(treatment))
print(mean(control))
diff <- mean(treatment)-mean(control)
print(diff)
```

So the hf diet mice are about 10% heavier. Are we done? Why do we need p-values and confidence intervals? The reason is that these averages are random variables. They can take many values. 

Note that we repeat the experiment, we will obtain 24 new mice from Jackson Laboratories and when we randomly assign them to each diet we will get a different mean. Everytime we repeat this experiment we get a different value. We call this type quantity a *random variable*. 

## Random variables

Let's see this. We actually have the weight of all control female mice and can load them up to R. In Stiatistcs we refer to this as *the population*. These are all the control mice avaiable from which we sampled 24.

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
population <- read.csv(filename)
```

Now let's sample 12 mice three times and see how the average changes.

```{r}
control <- sample(population[,1],12)
mean(control)

control <- sample(population[,1],12)
mean(control)

control <- sample(population[,1],12)
mean(control)
```

Note how the average varies. We can continue to do this over and over again and start learning something about 

## Null distribution

Now let's go back to our average difference of 2.23. A scientists we need to be skeptics. How do we know that this 2.23 is due to the diet. What happens if we give all 24 the same diet, can we see a difference this big? 

Because we have access to the population, we can actually observe as many values of the difference in average when there is no difference in diet. We can do this by sampling 12 control mice twice and taking their difference. Like this:

```{r}
control <- sample(population[,1],12)
treatment <- sample(population[,1],12)
print(mean(treatment) - mean(control))
```

Now let's do it 10,000 times. We will use a for-loop, an operation that let's us automatize this

```{r}
n <- 10000
null <- vector("numeric",n)
for(i in 1:n){
  control <- sample(population[,1],12)
  treatment <- sample(population[,1],12)
  null[i] <- mean(treatment) - mean(control)
}
```

The values in 'null' form what we call the *null distribution*. We will define this more formally later.

So what percent are bigger than 2.23?
```{r}
mean(null>=2.23)
```

Only 5.25%. So what do we conclude as skeptics. When there is no diet effect, we see value as big as 2.23 only 5.25% of the time. Note that this is what is known as a p-value which we will also define more formally.

##Summary
So this was pretty easy no? But why are we not done? Note that to make this calculation we bought all the mice available from Jackson labs and performed our experiment over and over again to define the null distribtuion. This is not something we can do in practice. Statistical Inference is the mathematical theory that permits you to approximate this with only the data from your sample, i.e. the original 24 mice. This is what we will learn in the chapter.



















## R system files
Note that this file is also included in the 'dagdata' package. If you have the package installed the this file is already in your system and you can use the 'system.file' function to find it

```{r}
dir <- system.file(package="dagdata")
list.files(dir)
list.files(file.path(dir,"extdata"))
filename <- file.path(dir,"extdata/miceweights.csv")
dat <- read.csv(filename)
```




If you installed the 'dagdata' package, then file we are using here is actually installed on your system. We included the file in the package for your use. You can find the file like this.








If you have installed the 'dagdat' package the file is already on your sytem here:

```{r}
system.file
```





the file. The path can be either to a file on your computer or to a file on the internet. 

The file we are reading in here can be obtaind both ways. 
We will read directly from the web using the 'read.csv' command

```{r}
file <- download.file("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/miceweights.csv",destfile="./miceweights.csv",method="curl")
dat <- read.csv(url(urlfile))
```






















In order to install Bioconductor, copy the following two lines into your R console.

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite()
```

This will install the core Bioconductor packages. Further packages can be installed using the `biocLite` function and specifying a character vector of which packages to install. For example, to install the "affy" and "genefilter" libraries you would type:

```{r, eval=FALSE}
biocLite(c("genefilter","geneplotter"))
```

Remember, you still need to load the library, e.g., `library(genefilter)`, if you want to use the package.

More information on installing and updating Bioconductor packages can be found at:

http://bioconductor.org/install/

## Finding help

There are many ways to find help directly from within R. Typically, every function will have its own manual page which is accessible by typing a question mark ?, followed by the function name and hitting return.

```{r, eval=FALSE, tidy=FALSE}
?mean
?mad
example(mad)
example(boxplot)
```

Simply typing the name of the function, without parentheses, and hitting return will show the source code of the function.

The manual page contains a **description**, example **usage**, explanation of all **arguments**, further **details**, explanation of the returned **value**, **references**, **see also** linking to other functions, and **examples**.

If you have a question about a particular object in R, you might want to look up the help for the "class" of that object, which will tell you how to construct such an object and what methods are available for manipulating such objects. For example, we can find the name of the class of an object and look up help:

```{r, eval=FALSE, tidy=FALSE}
class(6)
?numeric
?"numeric-class"
```

Sometimes, the constructor function and the class name will point to the same help page, although this is not necessarily true for all packages.

```{r, eval=FALSE, tidy=FALSE}
library(Biobase)
?ExpressionSet
?"ExpressionSet-class"
```

A quick way to find out what methods are available for a given class:

```{r, eval=FALSE}
methods(class="ExpressionSet")
methods(class="lm")
```

A quick way to look up functions in a given package is to write out the package name, two ":" symbols and then trying tab-completion to get a list of functions, exported or not.

```{r}
library(geneplotter)
```

```{r, eval=FALSE, tidy=FALSE}
geneplotter::
```

## Source code

You can find the source code for many functions by typing out the name of the function without () and pressing enter.

```{r}
mad
```

You might have to specify a particular class if you want source code for a method:

```{r}
plotMA
showMethods("plotMA")
getMethod("plotMA","data.frame")
```

## Vignettes

"Vignettes" are documents which accompany R packages and are required for every Bioconductor package. They typically show an example workflow of the functions of the package using "chunks" of code with descriptive text, exactly as the document you are currently reading. 

You can find Bioconductor vignettes in PDF or R script form on the Bioconductor website, but they are even easier to access directly through R. Furthermore, accessing vignettes through R guarantees that the vignette is for the correct version of the package that you are using. The following code will list the names of the vignettes for a given package:

```{r, eval=FALSE}
vignette(package="Biobase")
```

A further call to `vignette` with the name of the vignette will launch a PDF viewer:

```{r, eval=FALSE}
vignette("ExpressionSetIntroduction")
```

In addition, an HTML browser can be launched with links to the various vignettes of a package:


```{r, eval=FALSE, tidy=FALSE}
browseVignettes(package="Biobase")
```

