---
layout: page
title: Basic inference and EDA for high-throughput data
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r}
library(rafalib)
```

# Introduction 

Supposed we were given highthroughput gene expression data that was measured for several individuals in two populations. We are asked to report which genes have different average expression levels in the two populations. Note that if instead thousands of genes we were handed data from just one gene we could simply apply what 
we have learned before. We could, for example, use a t-test or some other test. Here we review what changes when we consider high-throughput data.

# Thousands of test

In this data we have two groups denoted with 0 and 1:
```{r}
library(GSE5859Subset)
data(GSE5859Subset)
g <- sampleInfo$group
g
```

If we were interested in a particular gene, say the 25th we would simply compute a t-test, assuming the data is well approximated by normal:

```{r}
e <- geneExpression[25,]
mypar2(1,2)
qqnorm(e[g==1])
qqline(e[g==1])
qqnorm(e[g==0])
qqline(e[g==0])
```

The t-test does not find this gene to be statistically singificant

```{r}
t.test(e[g==1],e[g==0])
```

To answer the question for each gene we simply to this for every gene

```{r}
myttest <- function(x) t.test(x[g==1],x[g==0],var.equal=TRUE)$p.value
pvals <- apply(geneExpression,1,myttest)
```

We can now see which genes have p-values less than, say, 0.05. For example right away we see that

```{r}
sum(pvals<0.05)
```

genes had p-values less than 0.05

However, as we will describe in more detail below, we have to be careful in the interpretation of this because we have performed over  8,000 test. If we did the same on random data:

```{r}
set.seed(1)
m <- nrow(geneExpression)
n <- ncol(geneExpression)
randomData <- matrix(rnorm(n*m),m,n)
nullpvals <- apply(randomData,1,myttest)
sum(nullpvals<0.05)
```

As we will explain this is to be expected. Note that 419 is roughly 0.05*8192. 

# Faster implementation of t-test

Before, we continue, we should note that the above implementation is very inneficiant. There are several faster implementations that perform t-test for high throughput data. For example

```{r}
library(genefilter)
results <- rowttests(geneExpression,factor(g))
max(abs(pvals-results$p))
```

# Volcano plots

As we described in the Introduction chapter, reporting only p-values is a mistake when we can aslo report effect sizes. With high-throughput data we can visualize the results by makine a plot. The idea behind a _volcano plot_ is to show these for all features. In the y-axis we plot -log (base 10) p-values and on the x-axis the effect size. By using - log (base 10) we have the "highly significant" results be high in the plot. Using log permits us to better distinguis between, say, 0.05 and 0.001.  Here is the volcano plot for our results above:

```{r}
plot(results$dm,-log10(results$p.value),xlab="Effect size",ylab="- log (base 10) p-values")
```


# p-value histograms

Another plot we can make to get an overall idea of what our results are telling us is to make histograms of p-values. Note that when we generate completely null data the histrogram follows a uniform distribtuion (we will say more about this). With our original data set we see a higher frequency of smaller p-values

```{r}
mypar2(1,2)
hist(nullpvals,ylim=c(0,1400))
hist(pvals,ylim=c(0,1400))
```



# Data boxplots and histogram

```{r}
mypar2(1,1)
boxplot(geneExpression)
```


```{r}
mypar2(1,1)
shist(geneExpression)
```


```{r}
quantiles <- t(apply(geneExpression,2,quantile,prob=c(0,0.05,0.25,0.5,0.75,0.95,1)))
matplot(quantiles,type="l")
```


# MA plot

```{r}
x <- geneExpression[,3]
y <- geneExpression[,16]
mypar(1,1)
plot(x,y)
plot((x+y)/2,x-y)
```










