---
layout: page
title: Multiple testing
---

The following code reproduces the images in the Multiple testing lecture.

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

First we make a function for drawing sections of a uniform distribution.

```{r}
drawU <- function(a,y,boxes=FALSE,xmax=1) {
  plot(1,1,type="n",xlim=c(0,xmax),ylim=c(0,1.2*y),bty="L",xlab="p",ylab="",las=1)
  lines(c(0,1,1,0,0),c(0,0,y,y,0))
  polygon(c(0,a,a,0),c(0,0,y,y),col=rgb(1,0,0,.5))
  polygon(c(a,1,1,a),c(0,0,y,y),col=rgb(0,0,0,.1))
  x <- 1/a
  if (boxes) {
    segments(0:x/x, 0, 0:x/x, y)
  }
}
```

Here we draw a uniform distribution, and show how many p-values we expect at different cutoffs.

```{r}
drawU(.3,1)
drawU(1/20,1)
drawU(1/20,20000/20,TRUE)
drawU(1/100,20000/100,TRUE)
drawU(1/1000,20000/1000,TRUE,.01)
```

The distribution of p-values using a z-score.

```{r}
z <- rnorm(100)
brks <- 0:20/20
hist(pnorm(z),col="grey",main="",xlab="p",breaks=brks)
```

The same as above, but with now many more z-scores. Now the distribution looks more uniform.

```{r}
z <- rnorm(10000)
hist(pnorm(z),col="grey",main="",xlab="p",breaks=brks)
```

What happens if we spike in 500 small z-scores?

```{r}
z <- c(rnorm(10000), rep(-3.72,500))
h <- hist(pnorm(z),col="grey",main="",xlab="p",breaks=brks)
abline(h=10000/20,col="red",lwd=4,lty=3)
abline(h=h$counts[1],col="blue",lwd=4,lty=3)
```

Now, making the bins in the histogram smaller, i.e. looking for a smaller p-value threshold:

```{r}
h <- hist(pnorm(z),col="grey",main="",xlab="p",breaks=0:50/50)
abline(h=10000/50,col="red",lwd=4,lty=3)
abline(h=h$counts[1],col="blue",lwd=4,lty=3)
```

Even smaller...

```{r}
h <- hist(pnorm(z),col="grey",main="",xlab="p",breaks=0:100/100)
abline(h=10000/100,col="red",lwd=4,lty=3)
abline(h=h$counts[1],col="blue",lwd=4,lty=3)
```

Even smaller...

```{r}
h <- hist(pnorm(z),col="grey",main="",xlab="p",breaks=0:10000/10000)
abline(h=10000/10000,col="red",lwd=4,lty=3)
abline(h=sum(h$counts[1]),col="blue",lwd=4,lty=3)
```

This visualizes the [Benjamini Hochberg method](#foot).

```{r}
set.seed(1)
pvals <- c(runif(90),runif(10,0,.001))
plot(sort(pvals),xlab="i",ylab="p-value",ylim=c(0,1))
abline(0, .05/length(pvals))
legend("top",expression(slope~alpha/m))
```

```{r}
plot(sort(pvals),xlab="i",ylab="p-value",ylim=c(0,.03),xlim=c(0,13))
abline(0, .05/length(pvals))
legend("top",expression(slope~alpha/m))
```

## Footnotes <a name="foot"></a>

Yoav Benjamini and Yosef Hochberg, "Controlling the False Discovery Rate: A Practical and Powerful Approach to Multiple Testing". Journal of the Royal Statistical Society. 1995.
<http://www.jstor.org/discover/10.2307/2346101>
