---
layout: page
title: Principal Components Analysis
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

# Introduction

Here we give a brief introduction to principal component analysis (PCA). We have already covered the singular value decomposition (SVD) which as we will see is very much related. 

# Example: Twin Heights

We used this simulated example to demonstrate a simple rotation. As we will see here this rotation is very much related to PCA.


```{r,fig.align="center",fig.width=4,fig.height=4}
library(rafalib)
library(MASS)
N = 100
mypar2(1,1)
set.seed(2)
Y=mvrnorm(N,c(0,0),matrix(c(1,0.95,0.95,1),2,2))
LIM=c(-1,1)*max(abs(Y))
colnames(Y) <- c("Twin 1 (standardized height)","Twin 2 (standardized height)")
plot(Y,xlab="Twin 1 (standardized height)",ylab="Twin 2 (standardized height)",xlim=LIM,ylim=LIM)
```

Let  $\mathbf{Y}$ be $N \times 2$ matrix representing our data. Suppose we are given the task of finding a  $2 \times 1$ vector $\mathbf{v}_1$ such that $\mathbf{v}_1^\top \mathbf{v}_1 = 1$
and it maximizes $$(\mathbf{Y}\mathbf{v}_1)^\top (\mathbf{Y}\mathbf{v}_1)$$. Note that this can be viewed as a projection of each row of $\mathbf{Y}$ into the subspace spanned by $\mathbf{v}$. So we are looking for a transformation in which the coordinates show high variability.

Let's try $\mathbf{v}=(1,0)^\top$. This projection gives us the height of twin 1 shown in orange below. The sum of squares is shown in the title.

```{r,fig.align="center",fig.width=4,fig.height=4}
mypar2(1,1)
plot(Y,xlim=LIM,ylim=LIM,main=paste("Sum of squares :",round(crossprod(Y[,1]),1)))
abline(h=0)
apply(Y,1,function(y) segments(y[1],0,y[1],y[2],lty=2))
points(Y[,1],rep(0,nrow(Y)),col=2,pch=16,cex=0.75)
```


Can we find a direction with higher variability? How about
$\mathbf{v} =\begin{pmatrix}1&-1\end{pmatrix}'$ ? This does not satisfy $\mathbf{v}^\top\mathbf{v}= 1$ so let's instead try
$\mathbf{v} =\begin{pmatrix}1/\sqrt{2}&-1/\sqrt{2}\end{pmatrix}'$ 

```{r,fig.align="center",fig.width=4,fig.height=4}
v <- matrix(c(1,-1)/sqrt(2),ncol=1)
w=Y%*%v
mypar2(1,1)
plot(Y,main=paste("Sum of squares:",round(crossprod(w),1)),xlim=LIM,ylim=LIM)
abline(h=0,lty=2)
abline(v=0,lty=2)
abline(0,-1,col=2)
Z = w%*%t(v)
for(i in seq(along=w))
  segments(Z[i,1],Z[i,2],Y[i,1],Y[i,2],lty=2)
points(Z,col=2,pch=16,cex=0.5)
```

This relates to the difference between twins which we know is small. The sum of squares confirms this.

Finally, let's try:
$\mathbf{v} =\begin{pmatrix}1/\sqrt{2}&1/\sqrt{2}\end{pmatrix}'$ 

```{r,fig.align="center",fig.width=4,fig.height=4}
v <- matrix(c(1,1)/sqrt(2),ncol=1)
w=Y%*%v
mypar2(1,1)
plot(Y,main=paste("Sum of squares:",round(crossprod(w),1)),xlim=c(-3,3),ylim=c(-3,3))
abline(h=0,lty=2)
abline(v=0,lty=2)
abline(0,1,col=2)
points(w%*%t(v),col=2,pch=16,cex=1)
Z = w%*%t(v)
for(i in seq(along=w))
  segments(Z[i,1],Z[i,2],Y[i,1],Y[i,2],lty=2)
points(Z,col=2,pch=16,cex=0.5)
```

This is a re-scaled average height which has higher sum of squares. It 
turns out that there is mathematical procedure for determining which $\mathbf{v}$ maximizes the sum of squares and that the SVD provides it for us.

# The Principal Components

The orthogonal vector that maximizes the sum of squares

$$(\mathbf{Yv}_1)^\top \mathbf{Yv}_1$$ 

$\mathbf{v}_1$ is referred to as the _first principal component_ (PC). Also referred as  _first eigenvector_, $\mathbf{Yv}_1$
are the projections or coordinates or eigenvalues 

Note that each row of of $\mathbf{Y}$ gets a coordinate


If we define residuals $$\mathbf{r} = \mathbf{Y} - \mathbf{Yv}_1 \mathbf{v}_1^\top$$

The second PC is the vector that 

$$ \mathbf{v}_2^\top \mathbf{v}_2=1$$

$$ \mathbf{v}_2^\top \mathbf{v}_1=0$$

and maximizes  $$(\mathbf{rv}_2)^\top \mathbf{rv}_2$$ 

When $Y$ is $N \times m$ we repeat to find 3rd, 4th, ..., m-th PCs

# Singular Value Decomposition

The SVD $\mathbf{Y=UDV}^\top$ gives the PCs in columns of $\mathbf{V}$. Note that the first PC is actually close to the last $\mathbf{v}$ we considered:

```{r}
s = svd(Y)
round(s$v * sqrt(2),1)
```


