---
layout: page
title: Installing Bioconductor and finding help
---

## Installing Bioconductor

In order to install Bioconductor, copy the following two lines into your R console.

```{r, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite()
```

This will install the core Bioconductor packages. Further packages can be installed using the `biocLite` function and specifying a character vector of which packages to install. For example, to install the "affy" and "genefilter" libraries you would type:

```{r, eval=FALSE}
biocLite(c("genefilter","geneplotter"))
```

Remember, you still need to load the library, e.g., `library(genefilter)`, if you want to use the package.

More information on installing and updating Bioconductor packages can be found at:

http://bioconductor.org/install/

## Finding help

There are many ways to find help directly from within R. Typically, every function will have its own manual page which is accessible by typing a question mark ?, followed by the function name and hitting return.

```{r, eval=FALSE, tidy=FALSE}
?mean
?mad
example(mad)
example(boxplot)
```

Simply typing the name of the function, without parentheses, and hitting return will show the source code of the function.

The manual page contains a **description**, example **usage**, explanation of all **arguments**, further **details**, explanation of the returned **value**, **references**, **see also** linking to other functions, and **examples**.

If you have a question about a particular object in R, you might want to look up the help for the "class" of that object, which will tell you how to construct such an object and what methods are available for manipulating such objects. For example, we can find the name of the class of an object and look up help:

```{r, eval=FALSE, tidy=FALSE}
class(6)
?numeric
?"numeric-class"
```

Sometimes, the constructor function and the class name will point to the same help page, although this is not necessarily true for all packages.

```{r, eval=FALSE, tidy=FALSE}
library(Biobase)
?ExpressionSet
?"ExpressionSet-class"
```

A quick way to find out what methods are available for a given class:

```{r, eval=FALSE}
methods(class="ExpressionSet")
methods(class="lm")
```

A quick way to look up functions in a given package is to write out the package name, two ":" symbols and then trying tab-completion to get a list of functions, exported or not.

```{r}
library(geneplotter)
```

```{r, eval=FALSE, tidy=FALSE}
geneplotter::
```

## Source code

You can find the source code for many functions by typing out the name of the function with () and pressing enter.

```{r}
mad
```

You might have to specify a particular class if you want source code for a method:

```{r}
plotMA
showMethods("plotMA")
getMethod("plotMA","data.frame")
```

## Vignettes

"Vignettes" are documents which accompany R packages and are required for every Bioconductor package. They typically show an example workflow of the functions of the package using "chunks" of code with descriptive text, exactly as the document you are currently reading. 

You can find Bioconductor vignettes in PDF or R script form on the Bioconductor website, but they are even easier to access directly through R. Furthermore, accessing vignettes through R guarantees that the vignette is for the correct version of the package that you are using. The following code will list the names of the vignettes for a given package:

```{r, eval=FALSE}
vignette(package="Biobase")
```

A further call to `vignette` with the name of the vignette will launch a PDF viewer:

```{r, eval=FALSE}
vignette("ExpressionSetIntroduction")
```

In addition, an HTML browser can be launched with links to the various vignettes of a package:


```{r, eval=FALSE, tidy=FALSE}
browseVignettes(package="Biobase")
```

