---
title: "Annotating phenotypes and molecular function"
author: "Vince"
date: "March 19, 2015"
output: html_document
layout: page
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

# The phenotype concept


- "Phenotype" is an extremely broad term
- In this course, it connotes low-dimensional representation of observable characteristics of an organism
- Representation can be numerical or categorical
    * units of measurement should be recorded
    * "codes" for categorical items should be clear

## Example: ExperimentalData package on COPD

We cross-tabulate gender and disease status for individuals in a study of chronic obstructive pulmonary disease

```{r dodim}
library(COPDSexualDimorphism.data)
data(lgrc.expr.meta)
with(expr.meta, table(gender, diagmaj))
```
## Continuous by categorical

Here's a boxplot of pack-years distributions, stratified by
gender and disease status.  The stratum labels become clumsy.

```{r lkbx, fig.width=8,fig.height=4.5,dpi=300,out.width="1920px",height="1080px",}
gd = with(expr.meta, factor(paste(gender,diagmaj)))
expr.meta$gd = gd
library(ggplot2)
ggplot(expr.meta, aes(x=gd, y=pkyrs)) + geom_boxplot()
#plot(pkyrs~gd, data=expr.meta)
```
## Phenotype carefully and record faithfully

- Validated questionnaires and protocols
- Standardized terminology, units
- Precise phenotypic characterization fosters more accurate mechanistic modeling
- Caveat: molecular "basis" suggests causal directionality, but phenotype and environment can influence molecular state

# Computing tools for inference on molecular mechanisms

- "Molecular basis" is likewise a broad notion
- Systematic terminologies exist to help clarify what is asserted in a given hypothesis or finding
- At the boundaries of scientific knowledge, disagreement is common and terminologies diverge
- Two examples: 
    * What is a gene?
    * What is a gene's function?
    
# Gene: A concrete computational definition

- ORMDL3 is a gene implicated in genome-wide association
studies as a factor in risk of asthma
- Here's a view of its "structure" according to human reference build hg19 (use ph525x::modPlot)
```{r domo,fig.height=4,fig.width=7}
library(ph525x)
modPlot("ORMDL3", collapse=FALSE, useGeneSym=FALSE)
```
* This will change with new reference build GRCh38

# Characterizing ORMDL3 functionality

```{r dohum}
library(Homo.sapiens)
orfunc = select(Homo.sapiens, key="ORMDL3", keytype="SYMBOL", 
   columns=c("GO", "TERM"))
orfunc[,c("ONTOLOGY", "TERM")]
```
- Gene Ontology standardizes terminology for biological processes, cellular components, and molecular functions

# Summary

* Phenotype characterization is challenging and frequently non-standard
* Tokens available for data analysis in R are fairly simple and are used in ad hoc ways to characterize sample phenotype and condition
* Reasoning about molecular processes underlying phenotype and disease states is intrinsically complex
* Standardized vocabularies and models exist and are available in Bioconductor, but limitations must be admitted

