---
layout: page
title: Reading in microarray data
---

```{r options, echo=FALSE}
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## Affymterix CEL files
Affymterix one color arrays
We start by reading in the sample information table. This is usually created by the person who performed the experiment. 

```{r}
library(affy)
basedir <- "celfiles"
tab <- read.delim(file.path(basedir,"sampleinfo.txt"),check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(basedir)
fns%in%tab[,1] ##check
ab <- ReadAffy(filenames=file.path(basedir,tab[,1]),phenoData=tab)
```

This creates an AffyBatch object which object contains infomration you need.
=======
The raw data files for this lab are in the `rawdata` repository, available here:

<https://github.com/genomicsclass/rawdata>

Click Download ZIP in order to download all the files, unzip this file which should result in a `rawdata-master` folder. 

First we save the initial working directory, so we can return to it.

```{r}
wd <- getwd()
```

Now we can start reading in the files:

```{r}
datadir <- "rawdata-master"
basedir <- paste0(datadir, "/celfiles")
setwd(basedir)
library(affy)
tab <- read.delim("sampleinfo.txt",check.names=FALSE,as.is=TRUE)
rownames(tab) <- tab$filenames
tab
fns <- list.celfiles()
fns
fns %in% tab[,1] ##check
ab <- ReadAffy(phenoData=tab)
```

This creates an AffyBatch object which object contains the information you need.
>>>>>>> upstream/master

```{r}
dim(pm(ab))#perfect match probe signal levels for each sample
dim(pData(ab))
#columns of probe level matrix match the rows of pData
rownames(pData(ab))
colnames(pm(ab))
annotation(ab)
```
Turn the probe level data into gene level data
Note, this object You can then preprocess RMA
```{r}
e <- rma(ab)
```

<<<<<<< HEAD
If you are not interested in probe level data you could can use this function
```{r}
e <- justRMA(filenames=tab[,1],celfile.path=basedir,phenoData=tab)
```

=======
Now we go back to the previous working directory.

```{r}
setwd(wd)
```

If you are not interested in probe level data you could can use this function

```{r}
setwd(basedir)
ejust <- justRMA(filenames=tab[,1],phenoData=tab)
dim(ejust)
```


>>>>>>> upstream/master
##Agilent data
Two color arrays
```{r}
library(limma)
library(rafalib)
<<<<<<< HEAD
basedir <- "agilent"
targets <- readTargets(file.path(basedir,"TargetBeta7.txt"))
RG <- read.maimages(targets$FileName, source="genepix",path=basedir)
#signals from red and green channel 
dim(RG$R)
dim(RG$G)

#instead of storing red and green information, use M-A tramsformation
MA <- MA.RG(RG,bc.method="none")
dim(MA$M)
dim(MA$A)
#quick check. MA plot for the first sample.
plot(MA$A[,1],MA$M[,1])
#check the real images for artifacts

mypar(1,1)
imageplot(MA$M[,2], RG$printer, zlim=c(-3,3))
#there are artificial patterns in the image
dev.off()
```


<<<<<<< HEAD
=======
Now we go back to the previous working directory.

```{r}
setwd(wd)
```
>>>>>>> upstream/master



## oligo
We can also use oligo to read affy arrays

```{r}
detach("package:affy")
library(oligo)
<<<<<<< HEAD
basedir <- "celfiles"
tab <- read.delim(file.path(basedir,"sampleinfo.txt"),check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(basedir,listGzipped=TRUE)
fns%in%tab[,1] ##check
pd <- as(tab, "AnnotatedDataFrame")
efs <- read.celfiles(filenames=file.path(basedir,tab[,1]),phenoData=pd,sampleNames=sampleNames(pd))
=======
basedir <- paste0(datadir,"/celfiles")
setwd(basedir)
tab <- read.delim("sampleinfo.txt",check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(listGzipped=TRUE)
fns %in% tab[,1] ##check
pd <- as(tab, "AnnotatedDataFrame")
efs <- read.celfiles(filenames=tab[,1],phenoData=pd,sampleNames=sampleNames(pd))
>>>>>>> upstream/master
```

```{r}
e <- rma(efs)
```
