# Reading in microarray data

## Affymterix CEL files

We start by reading in the sample information table. This is usually created by the person who performed the experiment. 

```{r}
library(affy)
basedir <- "celfiles"
tab <- read.delim(file.path(basedir,"sampleinfo.txt"),check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(basedir)
fns%in%tab[,1] ##check
ab <- ReadAffy(filenames=file.path(basedir,tab[,1]),phenoData=tab)
```

This creates an AffyBatch object which object contains infomration you need.

```{r}
dim(pm(ab))
dim(pData(ab))
annotation(ab)
```

Note, this object You can then preprocess RMA
```{r}
e <- rma(ab)
```

If you are not interested in probe level data you could can use this function
```{r}
e <- justRMA(filenames=tab[,1],celfile.path=basedir,phenoData=tab)
```

##Agilent data

```{r}
library(limma)
library(rafalib)
basedir <- "agilent"
targets <- readTargets(file.path(basedir,"TargetBeta7.txt"))
RG <- read.maimages(targets$FileName, source="genepix",path=basedir)
MA <- MA.RG(RG,bc.method="none")
mypar(1,1)
imageplot(MA$M[,2], RG$printer, zlim=c(-3,3))
dev.off()
```





## oligo
We can also use oligo to read affy arrays

```{r}
detach("package:affy")
library(oligo)
basedir <- "celfiles"
tab <- read.delim(file.path(basedir,"sampleinfo.txt"),check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(basedir,listGzipped=TRUE)
fns%in%tab[,1] ##check
pd <- as(tab, "AnnotatedDataFrame")
efs <- read.celfiles(filenames=file.path(basedir,tab[,1]),phenoData=pd,sampleNames=sampleNames(pd))
```

```{r}
e <- rma(efs)
```






