# Modeling

"All models are wrong, but some are useful" -George E. P. Box

When we see a p-value in the literature it means a probability distribution of some sort was used to quatify the null hypothesis. Many times deciding which probability distribution to use a relatively straight forward. For example, in the tea tasting example. Many p-values in the scientific literature are based on sample averages, or least squares estimates from a linear model, and make use of the CLT to approximate the null distribution of their statistic as normal.

The CLT is backed by theoretical results that guarantee that the approximation is accurate. However, we cannot always use this approximation, for example when our sample size is too small. In a previous module we described how the sample average can be approximated with t-distribution when the population data is approximately normal. Note that there is no theoretical backing for this assumption. We are now *modeling*. In the case of height, we know from experience that this turns out to be a very good model. 

```{r}
datadir="http://www.biostat.jhsph.edu/bstcourse/bio751/data"
dat=read.csv(file.path(datadir,"USheights_subsample.csv"))
men=dat$Height[dat$Gender==1]
qqnorm(men)
qqline(men)
```

But this doesn not imply that every dataset we collect will follow a normal distribtion. Example are: coin tosses, the numner of people who win the lottery, and US incomes.The normal is not the only parametric distribution that is available from modeling. Here we show we describe some useful parametric distribution and their use in genomics. For many more please consult these books [CITE books]

## The Binomial distribution

A distribution that one should be familiar is the binomial distribution. It described the probability of the total number of observed heads $S=k$ heads when tossing $N$ heads as
$$
\mbox{Pr}(S=k) = {N \choose k}p^k (1-p)^{N-k}
$$

with $p$ the probability of observing a head in one coin toss. Note that $S/N$ is the average of independent random variables and thus the CLT tells us that $S$ is approximately normal. This distribution is used by some of the variant callers and genotypers based on NGS to decide if the data is consistent 

## The Poisson distribution

The number of people that win the lottery follows a binomial distribution (we assume each person buys one ticket). The number of "tosses" $N$ is the number of people that buy tickets and very large. However, the number of people that win the lottery oscilates between 0 and 3. So why does CLT not hold? One can explain this mathematically, but the intuition is that with most the average so close to and also constrained to be larger than 0, it is impossible for the distribution to be normal. Here is a quick simulation

```{r}
p=10^-7 ##1 in 10,000,0000 chances of winning
N=5*10^6 ##5,000,000 tickets bought
winners=rbinom(1000,N,p) ##1000 is the number of different lotto draws
tab=table(winners)
plot(tab)
prop.table(tab)
```

For cases like this, where $N$ is very large but $p$ is small enough to make $N \times p$ (call it $\lambda$) a number between 0 and 10, then then $S$ can be shown to follow a Poisson a distribution which has a simple parametric form:
$$
\mbox{Pr}(S=k)=\frac{\lambda^k \exp{-\lambda}{k!}
$$

The Poisson distribution is commonly used in RNAseq analyses. Note that because we are sampling thousands moleculues and for some genes represent are a very small proportion of the totality of molecules, the Poisson distribution seems appropriate. 

Homework
1- Do you expect a Poisson distribution with $\lambda=100$ to be approximately nomral? Why or why not? 


So how does this help us? One way is that it informs us of the statistical properties of important summaries. For example, say we only have one sample from each of a case and control RNAseq experiment and we want to report the genes with larges fold-chages. Note that under the null, that there are no differences, the statistical variability of this quantity depends on the total abundance of the gene. We can show this mathematically but here is a quick simulation to demonstrate the point:
```{r}
N=10000##number of genes
lambdas=2^seq(1,16,len=N) ##these are the true abundances of genes
y=rpois(N,lambdas)##note that the null hypothesis is true for all genes
x=rpois(N,lambdas) 
ind=which(y>0 & x>0)##make sure no 0s due to ratio and log
plot(log2(lambdas[ind]),log2(y/x)[ind])
```
Note that for lower values of lambda there is much more variability and that if we were to report anything with a fold change of 2 or more the number of false positives would be quite high for low.



## NGS experiments and the Poisson distribution

```{r}
library(parathyroidSE)
data(parathyroidGenesSE)
```

This library contains SummarizedExperiment data, which will be discussed in a later lab. The important thing to know is that the SummarizedExperiment has a matrix of data, similar to the ExpressionSet, where each row is a genomic feature, and each column is a sample. For this dataset, the value in single cell in the matrix is count of reads which aligned to a given gene for a given sample.

```{r}
se <- parathyroidGenesSE
as.data.frame(colData(se))
```

A similar plot of technical replicates reveals that the behaviour predicted by the model is present in real data

```{r}
x <- assay(se)[,23]
y <- assay(se)[,24]
ind=which(y>0 & x>0)##make sure no 0s due to ratio and log
plot((log2(x[ind])+log2(y[ind]))/2,log(x[ind]/y[ind]))
```

When it comes to modeling, one limitation of the Poisson model is 
that it lacks flexibility when it comes to scale. Note that the Poisson only has one parameter which determins its mean $\lambda$ which is also its variance (standard deviation squared).

If we compute the standard deviations across four individuals it is quite a bit higher than what is predicted by a poisson model. Assuming most genes are differentially expressed across individuals, then if the Poisson model is appropriate their should be a linear relationship in this plot:

```{r}
library(rafalib)
library(matrixStats)
vars=rowVars(assay(se)[,c(2,8,16,21)]) ##we now these four are 4
means=rowMeans(assay(se)[,c(2,8,16,21)]) ##different individulsa
splot(means,vars,log="xy") ##plot a subset of data
abline(0,1,col=red)
```
Note that the variability plotted here includes biological variability which the motivation for the Poisson does not include. In a later module we learn about a the negative binomial distribution which combines the sampling variability of a Poisson and biological variability. The negative binomial permits has two parameters and permits more flexibility for count data. The Poisson is a special case of the negative bionomial distribution.

## Distributions for positive continuos values

In a previous module we learned that different genes vary differently accross biological replicates. In a latter module [advanced inference/differential expression] we will demonstrate that knowing this distribution can help improve downstream analysis. 

So can we model the distribution of these standard errors? Are they normal? Note that we are modeling the population standard errors so CLT does not apply. Here are some exploratory plots of the sample standard errors

```{r}
library(maPooling)
data(maPooling)
pd=pData(maPooling)
strain=factor(as.numeric(grepl("b",rownames(pd))))
pooled=which(rowSums(pd)==12 & strain==1)
techreps=exprs(maPooling[,pooled])
individuals=which(rowSums(pd)==1 & strain==1)
##remove replicates
individuals=individuals[-grep("tr",names(individuals))]
bioreps=exprs(maPooling)[,individuals]
###now compute the gene specific standard deviations
library(matrixStats)
techsds=rowSds(techreps)
biosds=rowSds(bioreps)
###now plot
library(rafalib)
mypar()
shist(biosds,unit=0.1,col=1,xlim=c(0,1.5))
shist(techsds,unit=0.1,col=2,add=TRUE)
```

First notice that the normal distribution is not appropriate here since the right tail is rather large. Also, because SDs are strictly positive there is a limitation to how symmetric this distribution can be.
A qqplot shows it very clearly
```{r}
qqnorm(biosds)
qqline(biosds)
```

There are parametric distributions that posses these properties (striclty positive and _heavy_ right tails). Two example sarethe _gamma_ and _F_ distributions. The desnity of the gamma distribution is defined by 
$$
f(x;\alpha,\beta)=\frac{\beta^\alpha x^{\alpha-1}\exp{-\beta x}}{\Gamma(\alpha)}
$$
Itis defined by two parameters $\alpha$ and $\beta$ that can, indirectly, control location and scale. They also control the shape of the distribution. For more on this distribution please refer [CITE BOOKs]. 

Two special cases of the gamma distribution are the chi-sqaured distribution, which we used earlier to analyze a 2x2 table and the exponential distribution which we will use later in this section. For chi-square we $\alpha=\nu/2$ and $\beta=2$ with $\nu$ the degrees of freedom. For exponential we have $\alpha=1$ and $\beta=\lambda$ the rate.

The F-distribution comes up in analysis of variance (ANOVA). It is also always positive and has large right tails. Two parameters control it's shape:
$$
f(x,d_1,d_2)=\frac{1}{B\left( \frac{d1}{2},\frac{d2}{2}\right)}
  \left(\frac{d_1}{d_2}\right)^{\frac{d_1}{2}}  
  x^{\frac{d_1}{2}-1}\left(1+\frac{d1}{d2}x\right)^{-\frac{d_1+d_2}{2}}
$$  
with $B$ the _beta function_


### Modeling precision

In a later module we will learn about empirical bayes approaches to improve estimates of variance. In these cases it is mathematically convinient (see Bayesian book) to model the precision $1/\sigma^2$. We will model that with a 


















