
Q 4.4.1  Elaborate the visualization of GM12878 ESRRA peaks as follows

seqlevels(GM12878, force=TRUE) = paste0("chr", 1:22)
autoplot(GM12878, layout="karyogram", aes(colour=log(peak)))

what is the difference between the bars colored in light
blue vs those colored in darker blue or black?

( ) light blue bars are located where ESSRA peaks are relatively low
( ) light blue bars are located where ESSRA peaks are relatively high
( ) light blue bars correspond to genomic locations with higher probability of ESRRA binding
( ) light blue bars correspond to genomic locations with lower probability of ESRRA binding

Q 4.4.2 An integrative visualization.  The following code combines
information on two cell lines and the measured peak values.

```{r fig=TRUE}
library(ERBS)
data(HepG2)
data(GM12878)
HepG2$cell = "HepG2"
GM12878$cell = "Bcell"
tot = c(GM12878, HepG2)
tot$peak10 = tot$peak/10 # copes with automatic scale of y axis
seqlevels(tot, force=TRUE) = paste0("chr", 1:22)
library(ggbio)
p = autoplot(seqinfo(tot))
p = p + layout_karyogram(tot, aes(fill=cell, colour=cell), geom="rect") + 
    scale_colour_manual(values = alpha(c("green", "red"), .1))
p + layout_karyogram(tot, aes(x=start, y=peak10), ylim=c(15,30), 
    geom="point", color="blue", size=.8)
```

Q 4.4.3 One chromosome is distinguished in this visualization by exhibiting a
unusually dense set of binding events.

Perform the following computations after computing `tot` as in the
previous problem.
```
stot = split(tot, as.character(seqnames(tot)))
w = sapply(stot, function(x) sum(width(x))) 
sort(w/seqlengths(tot)[names(w)])
```

Pick the chromosomes with the least and most dense binding patterns.

( ) chr1
( ) chr5
( ) chr13
( ) chr17
( ) chr21

Q 4.5.1  Use the erbsViz package as created
during the video to perform the following computation:

```{r do1}
library(erbsViz)
jdemo = juxta()
```

Q 4.5.1  What is the class of `jdemo`?
A Tracks
E class(jdemo)

Q 4.5.2  What slot of the jdemo object can be altered to change the
background color of the cell line label?
A label.bg.color
E getClass(class(jdemo))

Q 4.5.3  Modify the juxta function to allow selection of the
background color of the label and reinstall the package.
How many warnings are emitted by R CMD check for the revised package?
```

Q 4.6.1  Create Cen.ele10


Q 4.6.2  length(seqlevels(Cen.ele10))


Q 4.7.1 result of select count(*) from go_term

Q 4.7.2 
