---
layout: page
title: ExpressionSet and SummarizedExperiment
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

```{r message=FALSE}
library(Biobase)
library(GEOquery)
```

```{r}
geoq <- getGEO("GSE9514")
names(geoq)
e <- geoq[[1]]
```

### ExpressionSet

```{r}
dim(e)
exprs(e)[1:3,1:3]
dim(exprs(e))

pData(e)[1:3,1:6]
dim(pData(e))
names(pData(e))
pData(e)$characteristics_ch1

fData(e)[1:3,1:3]
dim(fData(e))
names(fData(e))
head(fData(e)$"Gene Symbol")
head(rownames(e))

experimentData(e)
annotation(e)
```

### Summarized Experiment

```{r message=FALSE}
library(parathyroidSE)
```


```{r}
data(parathyroidGenesSE)
se <- parathyroidGenesSE
se
```


```{r}
dim(se)
assay(se)[1:3,1:3]
dim(assay(se))

colData(se)[1:3,1:6]
dim(colData(se))
names(colData(se))
colData(se)$treatment

rowData(se)[1]
class(rowData(se))
length(rowData(se))
head(rownames(se))
metadata(rowData(se))

exptData(se)$MIAME
abstract(exptData(se)$MIAME)
```

## Footnotes

For more information about the `GenomicRanges` package, check out the PLOS Comp Bio paper, which the authors of GenomicRanges published:

http://www.ploscompbiol.org/article/info%3Adoi%2F10.1371%2Fjournal.pcbi.1003118

For more information on *SummarizedExperiment*:

http://www.nature.com/nmeth/journal/v12/n2/abs/nmeth.3252.html

Also the software vignettes have a lot of details about the functionality. Check out "An Introduction to Genomic Ranges Classes". All of the vignette PDFs are available here:

http://www.bioconductor.org/packages/release/bioc/html/GenomicRanges.html

