---
title: "The ExpressionSet container"
author: "Vince"
date: "March 19, 2015"
output: html_document
layout: page
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

# Overview

We'll work with the basic representation of expression experiments
in Bioconductor.  An example is in package Biobase.

```{r do1}
library(Biobase)
data(sample.ExpressionSet)
sample.ExpressionSet
```

We'll abbreviate the name:

```{r do2}
samp = sample.ExpressionSet
```

# Queries and extractors

```{r do3}
dim(samp)
exprs(samp)[1:5,1:6]  # extract expression values
pData(samp)    # extract sample level data
experimentData(samp)
abstract(samp)  # special accessor
```

Have a look at annotation package pmid2MIAME function to see
how to extract abstracts of papers from pubmed.  These can be
bound into ExpressionSets with experimentData().

# Matrix-like subscripting

We can use matrix-like syntax directly to restrict the
ExpressionSet, getting back a new ExpressionSet
```{r doex}
samp[1:4,3:20]
samp[, samp$sex=="Male"]
```


