---
layout: page
title: "A note on visualization options"
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

There are many interesting approaches to visualizing genome-scale data.
Two major packages in Bioconductor are Gviz and ggbio.  Both represent
significant efforts at bridging the gap between graphics facilities
and various genomic data structures.

ggbio's `autoplot` method can be very useful for broad overviews.
For a GRanges instance, each range for which data exists can be
depicted as a band on the chromosome.  The karyogram layout
gives a genome-wide view, but it can be important to control
the handling of extra-chromosomal sequence levels.

```{r getl,echo=FALSE,results="hide"}
library(ERBS)
library(GenomeInfoDb)
library(ggbio)
```
```{r lkd, fig=TRUE}
library(ERBS)
data(HepG2)
library(GenomeInfoDb)  # trim all but autosomal chroms
seqlevels(HepG2, force=TRUE) = paste0("chr", 1:22)
data(GM12878)
seqlevels(GM12878, force=TRUE) = paste0("chr", 1:22)
library(ggbio)
autoplot(HepG2, layout="karyogram", main="ESRRA binding on HepG2")
```

Notice that the title is not printed, currently a bug.

```{r lkm,fig=TRUE}
autoplot(GM12878, layout="karyogram", main="ESRRA binding on GM12878")
```
