---
layout: page
title: Mapping features to genes
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## Using Bioconductor annotation packages

This unit will focus on mapping features to genes, i.e., getting annotation information from one format to another. We start by loading in the `maPooling` dataset from previous lectures.

```{r}
# library(devtools)
# install_github("dagdata","genomicsclass")
library(dagdata)
library(Biobase)
data(maPooling)
e <- maPooling
head(rownames(e))
annotation(e)
```

The annotation for this ExpressionSet is *rae230a*. Many platforms will have database annotation packages already existing on Bioconductor. We can access these, first by installing, and then loading the library. We will use the `AnnotationDbi` package to query the information in the library.

While in this unit we will use a microarray annotation package as an example, the same commands can be used for an organism package, such as the homo sapiens annotation package `org.Hs.eg.db`, which let's one query from one kind of gene annotation to another.

```{r}
# biocLite(paste0(annotation(e),".db"))
library(rae230a.db)
# biocLite("AnnotationDbi")
library(AnnotationDbi)
```

Annotation packages have *columns*, some of which may be *keys*. You can query the database using a *key*, and ask for one or more *columns* in return. We will use the rownames of the ExpressionSet as keys.

```{r}
columns(rae230a.db)
keytypes(rae230a.db)
head(keys(rae230a.db, keytype="PROBEID"))
head(rownames(e))
```

The following `select` call will return the Entrez ID, ENSEMBL ID, and gene symbol for each Probe ID, which are the rownames of the ExpressionSet.

```{r}
res <- select(rae230a.db, keys=rownames(e),
              columns=c("ENTREZID","ENSEMBL","SYMBOL"), 
              keytype="PROBEID")
head(res)
idx <- match(rownames(e), res$PROBEID)
```

We need to align the `res` object so that we pull out, in order, one row for each row of the ExpressionSet.

```{r}
head(rownames(e))
head(res$PROBEID,7)
head(idx)
```

Here we add the new information to the `fData` of `e`. If there were already information in `fData`, we would have used `cbind` to add the new columns. Note here that, since we have a one-to-many mapping, the `match` function gave us the first match that it found. You could also collapse all possible matches of the Probe ID to the Genes using `split` and `paste` with the `collapse` argument. However, here we keep it simple and just take the first match in the `res` object.

```{r}
fData(e) <- res[idx,]
head(fData(e),10)
all.equal(fData(e)$PROBEID, rownames(e))
```

## Using Biomart

An alternate way to map from one annotation to another is using the `biomaRt` package. For more information on which Biomarts are available and how to access them, see the `biomaRt` vignette.

```{r}
# biocLite("biomaRt")
library(biomaRt)
# vignette("biomaRt")
m <- useMart( "ensembl", dataset = "rnorvegicus_gene_ensembl")
map <- getBM(mart = m,
             attributes = c("ensembl_gene_id", "entrezgene"), 
             filters = "ensembl_gene_id", 
             values = fData(e)$ENSEMBL)
head(map)
```

Finally, we need to align the new information with the old information using the `match` function as before, again picking the first match from a one-to-many mapping. We see that for the most part the new and the old Entrez IDs are the same, though some differences occur when we pick one from the one-to-many mappings that exist.


```{r}
idx <- match(fData(e)$ENSEMBL, map$ensembl_gene_id)
fData(e)$NEW_ENTREZID <- map$entrezgene[idx]
head(fData(e))
mean(fData(e)$ENTREZID == fData(e)$NEW_ENTREZID, na.rm=TRUE)
```

