---
layout: page
title: Reading in microarray data
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## Affymterix CEL files

We start by reading in the sample information table. This is usually created by the person who performed the experiment. 

The raw data files for this lab are in the `rawdata` repository, available here:

<https://github.com/genomicsclass/rawdata>

Click Download ZIP in order to download all the files, unzip this file which should result in a `rawdata-master` folder. 

First we save the initial working directory, so we can return to it.

```{r}
wd <- getwd()
```

Now we can start reading in the files:

```{r}
datadir <- "rawdata-master"
basedir <- paste0(datadir, "/celfiles")
setwd(basedir)
library(affy)
tab <- read.delim("sampleinfo.txt",check.names=FALSE,as.is=TRUE)
rownames(tab) <- tab$filenames
tab
fns <- list.celfiles()
fns
fns %in% tab[,1] ##check
ab <- ReadAffy(phenoData=tab)
```

This creates an AffyBatch object which object contains the information you need.

```{r}
dim(pm(ab))
dim(pData(ab))
annotation(ab)
```

Note, this object You can then preprocess RMA
```{r}
e <- rma(ab)
```

Now we go back to the previous working directory.

```{r}
setwd(wd)
```

If you are not interested in probe level data you could can use this function

```{r}
setwd(basedir)
ejust <- justRMA(filenames=tab[,1],phenoData=tab)
dim(ejust)
```


##Agilent data

```{r}
library(limma)
library(rafalib)
basedir <- paste0(datadir, "/agilent")
setwd(basedir)
targets <- readTargets("TargetBeta7.txt")
RG <- read.maimages(targets$FileName, source="genepix")
MA <- MA.RG(RG,bc.method="none")
mypar(1,1)
imageplot(MA$M[,2], RG$printer, zlim=c(-3,3))
dev.off()
```


Now we go back to the previous working directory.

```{r}
setwd(wd)
```



## oligo
We can also use oligo to read affy arrays

```{r}
detach("package:affy")
library(oligo)
basedir <- paste0(datadir,"/celfiles")
setwd(basedir)
tab <- read.delim("sampleinfo.txt",check.names=FALSE,as.is=TRUE)
fns <- list.celfiles(listGzipped=TRUE)
fns %in% tab[,1] ##check
pd <- as(tab, "AnnotatedDataFrame")
efs <- read.celfiles(filenames=tab[,1],phenoData=pd,sampleNames=sampleNames(pd))
```

```{r}
e <- rma(efs)
```
