---
layout: page
title: Surrogate variables and batch effects 
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## Confounding

Admission data from Berkeley 1973 showed more men were being admitted than women: 44\% men admitted compared to 30\% women!

```{r}
datadir="http://www.biostat.jhsph.edu/bstcourse/bio751/data"
dat=read.csv(file.path(datadir,"admissions.csv"))
dat$total=dat$Percent*dat$Number/100
##percent men get in
sum(dat$total[dat$Gender==1]/sum(dat$Number[dat$Gender==1]))
##percent women get in
sum(dat$total[dat$Gender==0]/sum(dat$Number[dat$Gender==0]))
```

A chi-square test clearly rejects they hypothesis that gender and admission are independent:
```{r}
##let's make a 2 x 2
indexes<-list(1:6,7:12)
tab <- matrix(NA,2,2,dimnames=list(c("males","females"),c("accepted","rejected")))
for(i in 1:2){
  ind<-indexes[[i]]
  tab[i,1]<- sum(dat[ind,2]*dat[ind,3]/100)
  tab[i,2] <- sum(dat[ind,2]*(1-dat[ind,3]/100))
}
tab <- round(tab)
tab
print(chisq.test(tab)$p.val)
```

But closer inspection shows a paradoxical results. Here are the percent admissions by Major:
```{r}
y=cbind(dat[1:6,c(1,3)],dat[7:12,3])
colnames(y)[2:3]=c("Male","Female")
y
```

#### *Homework*
1. Run the `chisq.test` in each major.

```{r}
y=cbind(dat[1:6,c(1,2)],dat[7:12,2])
colnames(y)[2:3]=c("Male","Female")
y
```

What's going? 

This is called _Simpson's paradox_ 
Note that males were much more likely to apply to "easy" majors. 

Male and easy majors are confounded. 
```{r}
y=cbind(dat[1:6,5],dat[7:12,5])
y=sweep(y,2,colSums(y),"/")*100
x=rowMeans(cbind(dat[1:6,3],dat[7:12,3]))
matplot(x,y,xlab="percent that gets in the major",ylab="percent that applies to major",col=c("blue","red"),cex=1.5)
legend("topleft",c("Male","Female"),col=c("blue","red"),pch=c("1","2"),box.lty=0)
```

```{r}
###make data for plot
library(rafalib)
mypar()
makematrix<-function(x,m,addx=0,addy=0){
  n<-ceiling(length(x)/m)
  expand.grid(1:n+addx,addy+1:m)[seq(along=x),] 
}
males<- sapply(1:6,function(i){
  tot=dat[i,2]
  p=dat[i,3]/100
  x=rep(c(0,1),round(tot*c(1-p,p)))
})
allmales<-Reduce(c,males)
females<- sapply(7:12,function(i){
  tot=dat[i,2]
  p=dat[i,3]/100
  rep(c(0,1),round(tot*c(1-p,p)))
})
allfemales<-Reduce(c,females)
plot(0,type="n",xlim=c(0,50),ylim=c(0,100),xaxt="n",yaxt="n",xlab="",ylab="")
PCH=LETTERS[rep(1:6,sapply(males,length))]
o<-order(allmales)
points(makematrix(allmales,100),col=2-allmales[o],pch=PCH[o],cex=0.6)
PCH=LETTERS[rep(1:6,sapply(females,length))]
o<-order(allfemales)
points(makematrix(allfemales,100,30),col=2-allfemales[o],pch=PCH[o],cex=0.6)
abline(v=29)
plot(0,type="n",xlim=c(0,80),ylim=c(0,130),xaxt="n",yaxt="n",xlab="",ylab="")
for(i in seq(along=males)){
  points(makematrix(males[[i]],20,0,22*(i-1)),col=2-sort(males[[i]]),pch=LETTERS[i],cex=0.6)
  points(makematrix(females[[i]],20,47,22*(i-1)),col=2-sort(females[[i]]),pch=LETTERS[i],cex=0.6)
  if(i>1) abline(h=22*(i-1)-0.5)
  }
abline(v=45)
```

So if we condition or stratify by major this goes away. 
```{r}
y=cbind(dat[1:6,3],dat[7:12,3])
matplot(1:6,y,xaxt="n",xlab="major",ylab="percent",col=c("blue","red"),cex=1.5)
legend("topright",c("Male","Female"),col=c("blue","red"),pch=c("1","2"),box.lty=0,cex=0.75)
```

The average difference by Major is 3.5% higher for women.

```{r}
mean(y[,1]-y[,2])
```

We see this in Baseball often:

|               | 1995           | 1996           | Combined        |
| ------------- | -------------- | -------------- | --------------- |
| Derek Jeter   | 12/48 (.250)   | 183/582 (.314) | 195/630 (.310)  |
| David Justice | 104/411 (.253) | 45/140 (.321)  | 149/551 (.270)  |

## Confounding in genomics

Here are the p-values from comparing caucasians to asians
```{r}
library(Biobase)
library(genefilter)
library(dagdata)
data(GSE5859)
eth<-factor(pData(e)$ethnicity=="CEU")
tt<-rowttests(exprs(e),eth)
HLIM<-c(0,6500)
mypar(1,2)
hist(tt$p.value,main="",xlab="p-values",nc=20,ylim=HLIM)
plot(tt$dm,-log10(tt$p.value),xlab="Effect size",ylab="-log10 (p-value)",xlim=c(-2,2))
```
Is this is really possible?

Note the confounding
```{r}
year=format(pData(e)[,2],"%Y")
table(year,pData(e)[,1])
```


##split by month
Let's compare the ASN from two June 2005 to October 2005
```{r}
batch <- format(pData(e)$date,"%y%m")
ind<-which(batch%in%c("0506","0510"))
tt<-rowttests(exprs(e)[,ind],factor(batch[ind]))
hist(tt$p.value,main="",xlab="p-values",nc=20,ylim=HLIM)
plot(tt$dm,-log10(tt$p.value),xlab="Effect size",ylab="-log10 (p-value)",xlim=c(-2,2))
```

