---
layout: page
title: Confidence Intervals
---

# Introduction

```{r,results=FALSE,echo=FALSE}
set.seed(1) ##so that we get same results
```

## Start by reading in the data
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
dat <- read.csv(filename)
```

```{r}
head(dat)
```

Let's start by selecting only female mice on the control diet

```{r}
chowPopulation <- dat[dat$Sex=="F" & dat$Diet=="chow",3]
```

and let's look at the population avarage which is our parameter of interest here:

```{r}
mu_chow <- mean(chowPopulation)
print(mu_chow)
```

We are interested in estimating this paramater. We want to know the average weight of females on the control diet but, in practice do not get to see the entire population. So we take a sample

```{r}
N <- 30
hf <- sample(chowPopulation,N)
```

We know this is a random variable the sample average will not be a perfect estimate. In fact, because in this illustrative example we know the value of the parameter, they are not exactly the same. A confidence interval is a statistical way of reporting our finding, the sample average, in a way that explicitely summarizes the variability of our random variable.

We will use the CLT. The CLT tells us that `mean(hf)` follows a normal distribution with mean `mean(chowPopulation)` and standard error approximately 

```{r}
se=sd(hf)/sqrt(N)
print(se)
```

A 95% condifence interval (we can use other percentages) is an random interval (start and end are random variables) with probability 95% of falling on the parameter. In a previon section we showed how we construct it. Let's do it here and see it in practice

```{r}
plot(mean(chowPopulation)+c(-7,7),c(1,1),type="n",xlab="weight",ylab="interval",ylim=c(1,100))
abline(v=mean(chowPopulation))
Q <- qnorm(1- 0.05/2)
lines( c(mean(hf)-Q*se, mean(hf)+Q*se ), c(1,1))
for(i in 2:100){
  hf <- sample(chowPopulation,N)
  se=sd(hf)/sqrt(N)
  interval <- c(mean(hf)-Q*se, mean(hf)+Q*se )
  color <- ifelse(mean(chowPopulation)<= interval[2] & mean(chowPopulation)>=interval[1],1,2)
  lines( interval, c(i,i),col=color)
}
```

You can take the code above and change N to see what happens. Note that if we make N too small, say 5, then the CLT no longer applies. So what do you expect to happen? Let's actually try it out?


```{r}
plot(mean(chowPopulation)+c(-7,7),c(1,1),type="n",xlab="weight",ylab="interval",ylim=c(1,100))
abline(v=mean(chowPopulation))
Q <- qnorm(1- 0.05/2)
N<-5
for(i in 1:100){
  hf <- sample(chowPopulation,N)
  se=sd(hf)/sqrt(N)
  interval <- c(mean(hf)-Q*se, mean(hf)+Q*se )
  color <- ifelse(mean(chowPopulation)<= interval[2] & mean(chowPopulation)>=interval[1],1,2)
  lines( interval, c(i,i),col=color)
}
```

Note that all the red. This is because the CLT is incorrectly telling us that the distribution of the `mean(hf)` is approximately normal when in fact it has fatter tail. The t-distribution might be more appropriate. All we have to do is change how we calculate `Q`: use `qt` instead of `qnorm`


```{r}
plot(mean(chowPopulation)+c(-7,7),c(1,1),type="n",xlab="weight",ylab="interval",ylim=c(1,100))
abline(v=mean(chowPopulation))
Q <- qt(1- 0.05/2, df=4)
N<-5
for(i in 1:100){
  hf <- sample(chowPopulation,N)
  se=sd(hf)/sqrt(N)
  interval <- c(mean(hf)-Q*se, mean(hf)+Q*se )
  color <- ifelse(mean(chowPopulation)<= interval[2] & mean(chowPopulation)>=interval[1],1,2)
  lines( interval, c(i,i),col=color)
}
```




## Connection between confidence intervals and p-values


```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
dat <- read.csv(filename)
```

We now have experimental data. This is no longer an illustration. This is what you will have in a typical data analysis

```{r}
control <- dat[1:12,2]
treatment <- dat[13:24,2]
diff <- mean(treatment)-mean(control)
print(diff)
```


Note that when we run the t-test function we actually get a 95% confidence interval (you can change the 0.95 through the arguments)
```{r}
t.test(treatment,control)
```

Note that if we provide a confidence interval in a way we also provide a p-value. Here is how. The null hypothesis is that the difference is 0. If a 95% confidence interval includes 0, this must mean that the p-value is larger than 0.05 because by definition of the confidence interval,  
`abs(mean(treatment)-mean(control))` is smaller than the quantile that gives you a p-value smaller than 0.05. Homework: show this mathematically.




