---
layout: page
title:  dplyr exercises
---

### Exercises

For these exercises we will use a new  dataset related to mammalian sleep. This data is described 
[here](http://docs.ggplot2.org/0.9.3.1/msleep.html). Download the CSV file from this location:

```{r ,echo=FALSE}
url="https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
filename <- basename(url)
```

We are going to read in this data, then test your knowledge of key `dplyr` functions `select` and `filter`. We are also going to review two different _classes_: data frames and vectors.

1. Read in the `msleep_ggplot2.csv` file with the function `read.csv` and use the function `class` to determine the what type of object is returned.

```{r answer1, echo=FALSE, eval=FALSE}
dat <- read.csv(filename)
class(dat)
```

2. Now use the `filter` function to select only the primates. How many animals in the table are primates? Hint: the `nrow` function gives you the number of rows of a data frame or matrix.

```{r answer2,echo=FALSE, eval=FALSE}
library(dplyr)
head(dat)
dat2 <- filter(dat, order=="Primates")
nrow(dat2)
```

3. What is the class of the object you obtain after subsetting the table to only include Primates?

```{r answer3,echo=FALSE, eval=FALSE}
dat2 <- filter(dat, order=="Primates")
class(dat2)
```

4. Now use the `select` function to extract the sleep (total) for the primates. What class is this object? Hint use `%>%` to pipe the results of the  `filter` function to `select`.

```{r answer4,echo=FALSE, eval=FALSE}
y <- filter(dat, order=="Primates") %>% select(sleep_total)
class(y)
```

5. Now we want to calculate the average of sleep for primates (the average of the numbers computed above). One challenge is the `mean` function requires a vector so if we simply apply it to the output above we get an error. Look at the help file for `unlist` and use it to compute the desidered average.

```{r answer5,echo=FALSE, eval=FALSE}
y <- filter(dat, order=="Primates") %>% select(sleep_total) %>% unlist
mean( y )
```

6. For the last exercise, we could also use the dplyr `summarize` function. We have not introduced this function but you can read the help file and repeat exercise 5 this time using just `filter` and `summarize` to get the answer.


```{r answer6,echo=FALSE, eval=FALSE}
filter(dat, order=="Primates") %>% summarise( mean( sleep_total) )
```

