---
layout: page
title: Getting Started
---

# Getting started

In this book we will be using the [R](https://cran.r-project.org/) programming language for all our analysis. You will learn R and statistics simultaneusly. However, we assume you have some basic programming skills and knowledge of R syntax. If you don't, the first homework, below, is to complete a tutorial. Here we give step-by-step instructions on how to get set up to follow along.


## Installing R

The first step is to install R. You can downloand and install R from the (Comprehensive R Archive Network)[https://cran.r-project.org/] (CRAN). It is relatively straight forward but if you need furthter help you can try the following resources:

* [Installing R on Windows](https://github.com/genomicsclass/windows#installing-r). 
* [Installing R on Mac](http://youtu.be/Icawuhf0Yqo)
* [Installing R on Ubuntu](http://cran.r-project.org/bin/linux/ubuntu/README)

## Installing RStudio

The next step is to install RStudio. Technically you can run all the code shown here without installing RStudio, but we highly recommend this integrated development environment (IDE). Instructions are [here](http://www.rstudio.com/products/rstudio/download/) and for Windows we have special [instructions](https://github.com/genomicsclass/windows).

## Learn R basics

The first homework assignment is to complete an R tutorial to familiarize yourself with the basics of programming and R syntax. If you are already familiar you can skip to the next section.

*Homework 0*:  Go through the [swirl](http://swirlstats.com/) tutorial, which teaches you R programming and data science interactively, at your own pace, and right in the R console”. Once you have R installed you can install swirl and run it this way:

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

Alternatively you can take the [try R](http://tryr.codeschool.com/) interactive class from Code School. 

Note that there are also many open and free resources and reference guides for R. Two examples are
* [Quick-R](http://www.statmethods.net/): a quick online reference for data input, basic statistics and plots
* R referecence card (PDF)[https://cran.r-project.org/doc/contrib/Short-refcard.pdf] by Tom Short 

Two key things you absolutely need to know about R is that you can get help for a function using `help` or `?`, like this:
```{r,eval=FALSE}
?install.packages
help("install.packages")
```

and that the hashtag symbol represents comments and are not interpreted:

```{r}
## This is just a comment
```


## Installing packages

The first R command we will run is `install.packages`. If you took the swirl tutorial you already did this! R only includes a basic set of functions. There is much more it can do than this, but not everybody needs everything so we instead make some functions via packages. Many of these function are stored in CRAN. Note that these packages are vetted. You can install easily from within R if you know the name of the function. As an example, we are going to install the package `downloader` which we use in our first data analysis examples:

```{r,eval=FALSE}
install.packages("downloader")
```

We can then load the package into our R sessions using:

```{r}
library(downloader)
```


From now on you will see that we sometimes load packages without installing them. This is because once you isntall it the pacakge remains in place and only needs to be loaded. If you try to load a package and get an error, the most likely reason for the error is that you need to install it.

## Importing data into R

The first step when getting ready to analyze data is to read in the data into R. There are several ways to do this and we will discuss three of them. But you only need to learn one to follow along. 

In the life sciences, small datasets such as the one used as an example in the next sections are stored as Excel file. Although there are R packages designed to read Excel (xls) format, in general you want to avoid the this and save files as comma delimited (csv) or tab delinted (txt) files. We will start with a simple example dataset containing [female mouse weights](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv). 

The first step is to find the file containing your data and know it's *path*. 


### Paths and the working directory

When you are working in R it is useful to know your _working directory_. This is the directory or folder in which R will save or look for files by default. You can see your working directory by typing:

```{r}
getwd()
```

You can also change your working directory using the function `setwd`. Or you can change it through RStudio by clicking on "Session". 

The functions that read and write files (there are several in R)
 assume you meant to look for files or write files in the working directory. Our recommended approach for beginners will have you  reading and writing to the working directory. But you can also type the [full path](http://www.computerhope.com/jargon/a/absopath.htm) which will work independently of the working directory.

### Projects in RStudio

We find that the simplest way to organize yourself is to start a Project in RStudio (Click on "File"" and the "New Project"). When creating the project, you will select a folder to be associated with it. You can then download all your data into this folder. Your working directory will be this folder.


### Option 1: Download file with your Browser to your working directory


You can navigate to the `femaleMiceWeights.csv` file by visiting the data directory of dagadat on [github](https://github.com/genomicsclass/dagdata/tree/master/inst/extdata). Note that if you navigate to the file, to assure that the downloaded file is in a csv file you need to click on *Raw* on the upper right hand corner of the data and then use your browser's "Save As" function. Note that some browsers add a suffix to your file name by default. You do not want this. You want your file to be called `femaleMiceWeights.csv`.

Once you have this file in your working directory then you can simply read it in like this:

```{r,eval=FALSE}
dat <- read.csv("femaleMiceWeights.csv")
```

Note that if you did not receive any message then you probably read in the file successfully.

### Option 2: Download from within R

We store many of the datasets used in this course on (GitHub)[https://github.com/genomicsclass/]. You can save these file directly from the internet to your computer using R. For example, in this example, we are using the `download.file` function in the `downloader` pacakge to download the file to a specific location and then read it in. We assign it a random name and a random directory using the function `tempfile` but you can change this so the file is saved a directory of your choosing.

```{r,message=FALSE}
library(downloader) ##use install.packages to install
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- tempfile() ## creates a random name for a temporary file
download(url, destfile=filename)
```

We can then read it in like this:
```{r}
dat <- read.csv(filename)
```


### Option 3: Installing packages from github (*Advanced*)

Many of the dataset we include in this book are avaialbe insider packages that are available from GitHub. The reason we use GitHub rather than CRAN is that on GitHub we do not have to vet packages and this gives us much more flexibility. 

To install packages from GitHub you will need to install the `devtools` package:

```{r}
install.packages("devtools")
```

Note to Windows users: To use devtools you will have to also install `Rtools`. In general you will need to install packages as administrator. One way to do this is to start R as administrator. If you do not have permission to do this, then it is a bit [more complicated](http://www.magesblog.com/2012/04/installing-r-packages-without-admin.html). 

Now we are ready to install from GitHub. Note that we now use a different function:

```{r}
library(devtools)
install_github("genomicsclass/dagdata")
```

The file we are working with is actually included in this package. So once you install the pacakge, the file is on your computer. However, finding it is somewhat advanced. Here are the lines of code:

```{r}
dir <- system.file(package="dagdata") ##This function extracts the location of package
list.files(dir)
list.files(file.path(dir,"extdata")) ## we know external data is in this directory
```

And now we are ready to read in the file:
```{r}
filename <- file.path(dir,"extdata/femaleMiceWeights.csv")
dat <- read.csv(filename)
```


## Brief introduction to dplyr

The learning curve for R sytax is slow. One of the harder aspects that requires getting used to is subsetting data tables. The `dplyr` packages brings these tasks closer to the language and we are therefore going to introduce two simple functions: one is used to subset and the other to select columns.

Take a look at the dataset we read in:
```{r}
dat <- read.csv(filename)
View(dat)
```

Note that there are two types of diets and that this is denoted in the first column. Note also that if we want just the weights, we only need the second column. So if we want to weights for mice on the `chow` diet we subset and filter like this:

```{r}
library(dplyr) 
chow <- filter(dat, Diet=="chow") ##keep only the ones with chow diet
View(chow)
```

And now we can select just the column with the values

```{r}
chowVals <- select(chow,Bodyweight)
chowVals
```

A nice feature of the dplyr function is that you can perform consectutive tasks of the form do this and then to that do this other thing using this character a "pipe". We use this symbol to say "then to that do this": `%>%`. So the above can be done in just one line:

```{r}
chowVals <- filter(dat, Diet=="chow") %>% select(Bodyweight)
```

Note that in the second task we now longer have to specific that object we are editing since it is whatever comes from the previous call.

So that you can see the way we would do this in R without dplyr this is the code:

```{r}
chowVals <- dat[ dat$Diet=="chow", colnames(dat)=="Bodyweight"]
```








