# Introduction

The unprecedented advance in digital technology during the second half
of the 20th century has produced a measurement revolution that is
transforming science. In the life sciences, data analysis is now part
of practically every research project. Genomics, in particular, is
being driven by new measurement technologies that permit us to observe
molecular entities for the first time and that are leading to
discoveries analogous to discovering the microorganism and many others
permitted by the invention of the microscope. Choice examples of these
technologies are microarrays and next generation sequencing
technology.

Scientific fields that have traditionally relied upon simple data
analysis techniques have been turned on their heads by these
technologies. For example, in the past, researchers would measure the
transcription levels of a single gene of interest. Today, it is
possible to measure all 20,000+ human genes at once.  Advances such as
these have brought about a shift from hypothesis to discovery-driven
research. However, interpreting information extracted from these
massive and complex datasets requires sophisticated statistical skills
as one can easily be fooled by patterns arising by chance. This has
greatly elevated the importance of statistics and data analysis in
this field.

## What Does This Book Cover?

This book will cover several of the statistical concepts and data
analytic skills needed to succeed in data-driven life science
research. We go from relatively basic concepts related to computing
p-values to advanced topics related to analyzing high-throughput data.

We start with one of the most important topics in statistics and in
the life sciences: statistical inference. Inference is the use of
probability to learn population features from data. A typical example
is deciphering if two groups (for example, cases versus controls) are
different on average. Specific topics covered include the t-test,
confidence intervals, association tests, Monte Carlo methods,
permutation tests and statistical power. We make use of approximations
made possible by mathematical theory, such as the Central Limit
Theorem, as well as techniques made possible by modern computing. We
will learn how to compute p-values and confidence intervals and
implement basic data analyses. Throughout the book we will describe
visualization techniques in the statistical computer language *R* that
are useful for exploring new data sets. For example, we will use these
to learn when to apply robust statistical techniques.

We will then move on to an introduction to linear models and matrix
algebra. We will explain why it is beneficial to use linear models to
analyze differences across groups, and why matrices are useful to
represent and implement linear models. We continue with a review of
matrix algebra, including matrix notation and how to multiply matrices
(both on paper and in R). We will then apply what we covered on matrix
algebra to linear models. We will learn how to fit linear models in R,
how to test the significance of differences, and how the standard
errors for differences are estimated. Furthermore, we will review some
practical issues with fitting linear models, including collinearity
and confounding. Finally, we will learn how to fit complex models,
including interaction terms, how to contrast multiple terms in R, and
the powerful technique which the functions in R actually use to
robustly fit linear models: the QR decomposition.

In the third part of the book we cover topics related to
high-dimensional data. Specifically, we describe multiple testing,
error rate controlling procedures, exploratory data analysis for
high-throughput data, p-value corrections and the false discovery
rate. From here we move on to covering statistical modeling. In
particular, we will discuss parametric distributions, including
binomial and gamma distributions. Next, we will cover maximum
likelihood estimation. Finally, we will discuss hierarchical models
and empirical Bayes techniques and how they are applied in genomics.

We then cover the concepts of distance and dimension reduction. We
will introduce the mathematical definition of distance and use this to
motivate the singular value decomposition (SVD) for dimension
reduction and multi-dimensional scaling. Once we learn this, we will
be ready to cover hierarchical and k-means clustering. We will follow
this with basic introduction to machine learning.

We end by learning about batch effects and how component and factor
analysis are used to deal with this challenge. In particular, we will
examine confounding, show examples of batch effects, make the
connection to factor analysis, and describe surrogate variable
analysis.

## How Is This Book Different?

While statistics textbooks focus on mathematics, this book focuses on
using a computer to perform data analysis. Instead of explaining the
mathematics and theory and then showing examples, we start by stating
a practical data-related challenge. We then provide computer code that
provides a solution to the problem and also helps illustrate the
concepts behind the solution. By running the code yourself, and seeing
data generation and analysis happen live, you will get a better
intuition for the concepts, the mathematics, and the theory.

We focus on the practical challenges faced by data analysts in the
life sciences and introduce mathematics as a tool that can help us
achieve scientific goals. Furthermore, throughout the book we show the
R code that performs this analysis and connect the lines of code to
the statistical and mathematical concepts we explain. All sections of
this book are reproducible as they were produced by *R markdown*
documents that include R code used to produce the figures, tables and
results shown in the book. In order to distinguish it, the code is
shown in the following font:

```{r,eval=FALSE} x <- 2 y <- 3 print(x+y) ```

and the results in different colors and preceded by two hash
characters (*##*):

```{r,echo=FALSE} x <- 2 y <- 3 print(x+y) ```

We will provide links that will give you access to the raw R markdown
code so you can easily follow along with the book by programming in R.
