---
layout: page
title: Introduction to random variables and reading data
---

# Introduction

```{r,results=FALSE,echo=FALSE}
set.seed(1) ##so that we get same results
```

This course introduces the statistical concepts necessary to underatadn p-value and confidence intervals that are ubiquoitous in the life science literature. Let's take an example. This paper [http://diabetes.diabetesjournals.org/content/53/suppl_3/S215.full] states that 

"Body weight was higher in mice fed the high-fat diet already after the first week, due to higher dietary intake in combination with lower metabolic efficiency. "

To back this up they provide this

"Already during the first week after introduction of high-fat diet, body weight increased significantly more in the high-fat diet–fed mice (+1.6 ± 0.1 g) than in the normal diet–fed mice (+0.2 $\pm$ 0.1 g; P < 0.001)."

What does P < 0.001 mean? What are $\pm$ included? In this class we will learn what this mean and learn to compute these values in R. The first step is to understand what is a random variable. To understand this, we will use data from a mouse database [CITE: provided by Karen Svenson via Gary Churchill and <Gary.Churchill@jax.org> Dan Gatti and Partially funded by P50 GM070683.]. We will import the data with R and explain random variables and null distributions using R programming.

## Importing data into R
Before defining random variable and null distribution let's import data into R.

In the life sciences, small datasets such as the one used as an example here are stored as Excel file. In general you want to avoid the Excel (xls) format and save files as comma delimited (csv) or tab delimted (txt) files. The first step is to find this file and know it's *path*. 

### Finding the file

The file we are using can be found our github repository [here](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/miceweights.csv). Now once you download the file, how do you find the path? Windows and Macs have slighlty different ways of finding these. I'll leave to you to figure this out. 

You can also use R to downloade the file and define the path yourself. Let's download the file and put it in our Desktop. In general we don't recommend downloading files to your Desktop. We use it as an example.

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
```

Now we are ready to read in the file 
```{r}
dat <- read.csv(filename)
```

## Our first look at data

We are interesting determing if following a given diet makes mice heavier after several. This data was produced by ordering 24 mice from Jackson Lab, randomely assigning either chow or high fat (hf) diet. Then after several weeks we weighed each mice and obtained this data:

```{r}
dat
```

So are the hf mice heavier? Note that mouse 21 is 20.73 grams is one the lightest mice while 14 at 34.02 is one of the heaviest. Both are on the hf diet. Just from looking the data we see there is *variability*. Claims such as the one above usually refer to the averages. So let's look at the average of each grou

```{r}
control <- dat[1:12,2]
treatment <- dat[13:14,2]
print(mean(treatment))
print(mean(control))
diff <- mean(treatment)-mean(control)
print(diff)
```

So the hf diet mice are about 10% heavier. Are we done? Why do we need p-values and confidence intervals? The reason is that these averages are random variables. They can take many values. 

Note that we repeat the experiment, we will obtain 24 new mice from Jackson Laboratories and when we randomly assign them to each diet we will get a different mean. Everytime we repeat this experiment we get a different value. We call this type quantity a *random variable*. 

## Random variables

Let's see this. We actually have the weight of all control female mice and can load them up to R. In Stiatistcs we refer to this as *the population*. These are all the control mice avaiable from which we sampled 24.

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- tempfile()
download.file(url,destfile=filename,method="curl")
population <- read.csv(filename)
```

Now let's sample 12 mice three times and see how the average changes.

```{r}
control <- sample(population[,1],12)
mean(control)

control <- sample(population[,1],12)
mean(control)

control <- sample(population[,1],12)
mean(control)
```

Note how the average varies. We can continue to do this over and over again and start learning something about 

## Null distribution

Now let's go back to our average difference of 2.23. A scientists we need to be skeptics. How do we know that this 2.23 is due to the diet. What happens if we give all 24 the same diet, can we see a difference this big? Statisticians refered to this scenario as the *null hypothesis*. The name null is used to remind us that we are acting as skeptics: there is no difference. 

Because we have access to the population, we can actually observe as many values of the difference in average when the diet has no effect. We can do this by sampling 24 control mice, giving them the same diet, and then recording the difference in mean between to randomely split groups. Here is the code:

```{r}
control <- sample(population[,1],12)
treatment <- sample(population[,1],12)
print(mean(treatment) - mean(control))
```

Now let's do it 10,000 times. We will use a for-loop, an operation that let's us automatize this

```{r}
n <- 10000
null <- vector("numeric",n)
for(i in 1:n){
  control <- sample(population[,1],12)
  treatment <- sample(population[,1],12)
  null[i] <- mean(treatment) - mean(control)
}
```

The values in 'null' form what we call the *null distribution*. We will define this more formally below.

So what percent are bigger than 2.23?
```{r}
mean(null>=2.23)
```

Only 5.25%. So what do we conclude as skeptics. When there is no diet effect, we see value as big as 2.23 only 5.25% of the time. Note that this is what is known as a p-value which we will also define more formally.

##Interactive plot
Let's repeat the loop above but this time let's add a point to the figure everytime we re-run the experimetn

```{r}
n <- 100
plot(0,0,xlim=c(-5,5),ylim=c(1,30),type="n")
totals <- vector("numeric",11)
for(i in 1:n){
  control <- sample(population[,1],12)
  treatment <- sample(population[,1],12)
  nulldiff <- mean(treatment) - mean(control)
  j <- pmax(pmin(round(nulldiff)+6,11),1)
  totals[j]<-totals[j]+1
  text(j-6,totals[j],pch=15,round(nulldiff,1))
  if(i < 15) scan()
  }
```



## Distributions

We have explained what we mean by *null* in the context of null hypothesis but what exactly is a distribution?
The simplest way to think of a *distribution* is as a compact description of many numbers. For example, in the previous section we defined an object 'null' with 10,000 average differences created under the null. To define a distribution we compute, for all possible values of $a$ the proportion of numbers in our list that are below $a$. We use the following notation

$$ F(a) \equiv \mbox{Pr}(x \leq a) $$

This is called the empirical commulative distribution function. We can plot $F(a)$ versus $a$ like this

```{r}
values <- seq(min(null),max(null),len=300)
myecdf <- ecdf(null)
plot(values,myecdf(values),type="l")
```

These ecdfs are actually not as popular as histograms which give us the same information but show us the proportion of values in intervals

$\mbox{Pr}(a \leq x \leq b) = F(b)-F(a)$.

This is a more useful plot because we are usually more interested in intervals. It is also easier to distinguish different types (families) of distributions by looking at histograms. 

Note that from the histogram we can see that values as large as 2.23 are relatively rare
```{r}
hist(null)
abline(v=diff)
```

We will provide more details on histograms in later chapters. 

An important point to keep in mind here is that while we defined $Pr(a)$ by counting cases, we will learn how, in some circumstances, mathematics gives us formulas for $Pr(a)$ that save us the trouble of computing them as we did here.

##Summary
So this was pretty easy no? But why are we not done? Note that to make this calculation we bought all the mice available from Jackson labs and performed our experiment over and over again to define the null distribtuion. This is not something we can do in practice. Statistical Inference is the mathematical theory that permits you to approximate this with only the data from your sample, i.e. the original 24 mice. This is what we will learn in the following sections. 






















## R system files
Note that this file is also included in the 'dagdata' package. If you have the package installed the this file is already in your system and you can use the 'system.file' function to find it

```{r}
dir <- system.file(package="dagdata")
list.files(dir)
list.files(file.path(dir,"extdata"))
filename <- file.path(dir,"extdata/miceweights.csv")
dat <- read.csv(filename)
```



