---
layout: page
title: Rank tests
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

# Wilcoxon signed-rank test

We learned how the sample mean and sd  are susceptible to outliers. The t-test is based on these measures and is susceptible as well. The Wilcox signed-rank test provides and alternative. In the code below we perform a t-test on data for which the null is true but we change one observation by mistakes in each sample and the values incorrectly entered are different. Here we see that the t-test results in a small p-value while the Wilcoxon test does not.

```{r}
set.seed(779) ##for illustration
N=25
x<- rnorm(N,0,1)
y<- rnorm(N,0,1)
x[1]<-5;x[2]<-7
cat("t-test pval:",t.test(x,y)$p.value)
cat("Wilcox test pval:",wilcox.test(x,y)$p.value)
```

The basic idea is to 1) combine all the data, 2) turn the values into ranks, 3) separate them back into their groups, and 4) compute the sum or average rank and perform a test.


```{r}
stripchart(list(x,y),vertical=TRUE,ylim=c(-7,7),ylab="Observations",pch=21,bg=1,cex=1.25)
abline(h=0)
xrank<-rank(c(x,y))[seq(along=x)]
yrank<-rank(c(x,y))[-seq(along=y)]
stripchart(list(xrank,yrank),vertical=TRUE,ylab="Ranks",pch=21,bg=1,cex=1.25)
ws <- sapply(x,function(z) rank(c(z,y))[1]-1)
text(rep(1.05,length(ws)),xrank,ws)
W <-sum(ws) 
```

`W` is the sum of the ranks for the first group. Statistical theory (based on CLT) tells us that this statistic is approximated by the normal distribution. In fact we can construct a z-score as follows:

```{r}
n1<-length(x);n2<-length(y)
Z <- (mean(ws)-n2/2)/ sqrt(n2*(n1+n2+1)/12/n1)
print(Z)
```

Note that the `Z` is not large enough to give us a p-value less than 0.

