---
title: "t-tests in Practice"
layout: page
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.path=paste0("figure/", sub("(.*).Rmd","\\1",basename(knitr:::knit_concord$get('infile'))), "-"))
```

## t-tests In Practice

### Introduction

We will now demonstrate how to obtain a p-value in practice. We begin by loading experimental data and walking you through the steps used to form a t-statistics and compute a p-value. We can perform this task with just a few lines of code (go to end of section to see them). However, to understand the concepts, we will construct a t-statistic from "scratch".


### Setting The Random Seed
Before we continue we briefly explain the following important line of code:
```{r,results=FALSE,echo=FALSE}
set.seed(1) 
```

Throughout this book we use random number generators. This implies that many of the results presented can actually change by chance, including the correct answer to problems. One way to assure that results do not change is by setting the random number generation seed. For more on the topic please read the help file:

```{r,eval=FALSE}
?sed.seet
```



### Read-in and Prepare Data
We start by reading in the data. A first important step is to identify which rows are associated with treatment and control, and computing the difference in mean.
```{r,message=FALSE}
library(dplyr)

dat <- read.csv("femaleMiceWeights.csv") #previously downloaded

control <- filter(dat,Diet=="chow") %>% select(Bodyweight) %>% unlist
treatment <- filter(dat,Diet=="hf") %>% select(Bodyweight) %>% unlist

diff <- mean(treatment)-mean(control)
print(diff)
```

We are asked to report a p-value. What do we do? We learned that `diff`, referred to as the _observed effect size_, is a random variable. Under the null hypothesis what is the distribution of this random variable? Let's use what we learned.

Under the null, the mean of the distribution of `diff` is 0. What about the standard deviation? 

To simplify, let's start with `mean(control)`. This is also a random variable. We want to know the standard error of the distribution of this random variable, which from now on we will call a _standard error_ (SE). In statistics we call the standard deviation of the distribution of a random variable, the standard error of the random variable. Previously, we learned that statistical theory tells us that the standard error of this random variable is the population standard deviation divided by the square root of the square root of the sample size. The formula we showed was:

$$ SE(\bar{X}) = \sigma / \sqrt{N}$$

We use the sample standard deviation as an estimate of the SE. In R we simply use the `sd` function and the SE is simply:

```{r}
sd(control)/sqrt(length(control))
```

This is the SE of the sample average, but we actually want the SE of `diff`. We saw how statistical theory tells us that the variance of the difference of two random variables is the sum of its variances, so we compute the variance and take the square root:

```{r}
se <- sqrt( 
  var(treatment)/length(treatment) + 
    var(control)/length(control) 
  )
```

Statistical theory tells us that if we divide a random variable by its SE, we get a new random variable with SE of 1.

```{r}
tstat <- diff/se 
```

This ratio is what we call the t-statistics. It's the ratio of two random variables and thus a random variable. Once we know the distribution of this random variable, we can then easily compute a p-value.

As explained in the previous section, the CLT tells us that for large sample sizes both sample averages `mean(treatment)` and `mean(control)` are normal. Statistical theory tells us that the difference of two normally distributed random variables is again normal, so CLT tells us that `tstat` is approximately normal with mean 0 (the null hypothesis) and SD 1 (we divided by its SE). 

So now to calculate a p-value all we do is ask: how often does a normally distributed random variable exceed `diff`? R has a function specifically built in to answer this question: `pnorm`. `pnorm(a)` returns the probability that random variable following the standard normal distribution falls below `a`. To obtain the probability that it is larger than `a` we simply use `1-pnorm(a)`. We want to know the probability of seeing something as extreme as `diff`: either smaller (more negative) than `-abs(diff)` or larger than `abs(diff)`:

```{r}
righttail <- 1-pnorm(abs(tstat)) 
lefttail <- pnorm(-abs(tstat))
pval <- lefttail + righttail
print(pval)
```

In this case the p-value is smaller than 0.05 and using the conventional cutoff of 0.05, we would call the difference _statistically significant_.

Now there is a problem. CLT works for large samples, but is 12 large enough? A rule of thumb for CLT is that 30 is a large enough sample size (but this is just a rule of thumb). The p-value we computed is only a valid approximation if the assumptions hold, which does not seem to be the case here. However, there is another option other than using CLT.

<a name="smallsample"></a>

## The t-distribution in Practice

As described earlier, the statistical theory offers another useful result. If the distribution of the population is normal, then we can work out the exact distribution of the t-statistic without the need for the CLT. Note that this is a big "if" given that, with small samples, it is hard to check if the population is normal. But for something like weight, we suspect that the population distribution is likely well approximated by normal and use this approximation. Furthermore, we can look at qq-plot for the sample. This shows that the approximation is at least close:

```{r data_qqplot, fig.cap="Quantile-quantile plots for sample against theoretical normal distribution.", fig.width=10.5,fig.height=5.25}
library(rafalib)
mypar(1,2)

qqnorm(treatment)
qqline(treatment,col=2)

qqnorm(control)
qqline(control,col=2)
```

If we use this approximation, then statistical theory tells us that the distribution of the random variable `tstat` follows a t-distribution. This is a much more complicated distribution than the normal that depends on another parameter called degrees of freedom. R has a nice function that actually computes everything for us.

```{r}
t.test(treatment,control)
```

To see just the p-value we can use the `$` extractor:

```{r}
result <- t.test(treatment,control)
result$p.value
```


The p-value is slightly bigger now. This is to be expected because the CLT approximation considers the denominator of t-stat practically fixed (with large samples it practically is), while the t-distribution approximation takes into account that it is a random variable and that the smaller the sample size, the more it varies.

It may be confusing that one approximation gave us one p-value and another gave us another because we expect there to be just one answer. But this is not uncommon in data analysis. We used different assumptions, different approximations, and therefore we obtained different results.

Later, in the power calculation section, we will describe type I and type II errors. As a preview, we will point out that the test based on the CLT approximation is more likely to incorrectly reject the null (false positive), while the t-distribution is more likely to incorrectly accept the null (false negative).

### Running the t-test in Practice

Now that we have gone over the concepts, we can show the relatively simple code that one actually would run to compute a t-test: 


```{r,message=FALSE,echo=FALSE}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- "mice_pheno.csv"
if(!file.exists(filename)) download(url,destfile=filename)
```


```{r,message=FALSE}
library(dplyr)

dat <- read.csv("mice_pheno.csv")

control <- filter(dat,Diet=="chow") %>% select(Bodyweight) 

treatment <- filter(dat,Diet=="hf") %>% select(Bodyweight) 

t.test(treatment,control)
```

The arguments to `t.test` can be data.frames and thus we do not need to unlist them into numeric objects.
