# ChIP-Seq

```{r}
biocLite("DiffBind")
library(DiffBind)
setwd(system.file("extra", package="DiffBind"))
```

```{r}
list.files()
read.csv("tamoxifen.csv")
list.files("peaks")
```

```{r}
ta <- dba(sampleSheet="tamoxifen.csv")
ta
```

From the `DiffBind` vignette, we have:

"This shows how many peaks are in each peakset, as well as (in the first line) total number of unique peaks *after merging overlapping ones* (3557) and the default binding matrix of 11 samples by the 2602 sites that *overlap in at least two of the samples*."

```{r}
names(ta)
class(ta$peaks)
head(ta$peaks[[1]])
```

```{r}
ta <- dba.count(ta, minOverlap=3)
data(tamoxifen_counts)
ta2 <- tamoxifen
plot(ta2)
```

```{r}
ta2 <- dba.contrast(ta2, categories=DBA_CONDITION)
ta2 <- dba.analyze(ta2)
ta2
```

"By default, dba.analyze plots a correlation heatmap if it finds any significantly differentially bound sites, shown in Figure 3. Using only the differentially bound sites, we now see that the four tamoxifen resistant samples (representing two cell lines) cluster together, although the tamoxifen-responsive MCF7 replicates cluster closer to them than to the other tamoxifen responsive samples."


```{r}
tadb <- dba.report(ta2)
tadb
counts <- dba.report(ta2, bCounts=TRUE)
```

```{r}
x <- mcols(counts)[1,-c(1:6)]
x <- unlist(x)
(xord <- x[match(ta2$samples$SampleID, names(x))])
ta2$samples$SampleID
```

```{r}
cond <- factor(ta2$samples[,"Condition"])
condcomb <- factor(paste(ta2$samples[,"Condition"], ta2$samples[,"Tissue"]))
```

```{r}
par(mar=c(15,5,2,2))
stripchart(log(xord) ~ condcomb, method="jitter", 
           vertical=TRUE, las=2, ylab="log2 normalized counts")
means <- tapply(xord, cond, mean)
log2(means)
log2(means[1] / means[2])
mcols(tadb)$Fold[1]
```

